---
 title: "MCMC diagnostics - survival"
 subtitle: "Model: intercept + plot random effects  + competition"
 author: "Will Vieira"
 date: "`r paste('Last updated on', format(Sys.time(), '%d %B, %Y'))`"
---

```{r, echo = F, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(cmdstanr))
suppressPackageStartupMessages(library(loo))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(tidybayes))
suppressPackageStartupMessages(library(bayesplot))
suppressPackageStartupMessages(library(ggridges))
suppressPackageStartupMessages(require(RColorBrewer))
suppressPackageStartupMessages(library(DT))
suppressPackageStartupMessages(library(ggrepel))
```
```{r load simulations, echo = F,warning=FALSE,message=FALSE}
# Load simulation variables
model_dir <- 'model_survival/intcpt_plot_comp'
simInfo <- yaml::read_yaml(file.path(model_dir, '_simulation_info.yml'))
output_dir <- file.path('output_sim_processed', simInfo$vitalRates, simInfo$simName)

sp_info <- read_csv(file.path('..', 'forest-IPM', 'data', 'species_id.csv'))
spIds <- simInfo$spIds
simName <- simInfo$simName

# List output files and set vector names to species_id
posteriorPop_files <- grep(
  'posterior_pop',
  dir(output_dir, full.names = TRUE),
  value = TRUE
)
names(posteriorPop_files) <- names(sort(sapply(spIds, function(x) grep(x, posteriorPop_files))))

posteriormPlot_files <- grep(
  'posterior_plot',
  dir(output_dir, full.names = TRUE),
  value = TRUE
)
names(posteriormPlot_files) <- names(sort(sapply(spIds, function(x) grep(x, posteriormPlot_files))))

diag_files <- grep(
  'diagnostics',
  dir(output_dir, full.names = TRUE),
  value = TRUE
)
names(diag_files) <- names(sort(sapply(spIds, function(x) grep(x, diag_files))))
```


```{r load dataset,echo =FALSE,warning=FALSE,message=FALSE}
dataSource <- readRDS(
  gsub('../../', '', simInfo$dataSources)
) |>
filter(species_id %in% spIds)
```

# Rhat

```{r, echo = F,fig.height = 7, fig.width = 7,warning=FALSE,message=FALSE}
imap_dfr(
    diag_files,
    ~ readRDS(.x)[['rhat']] |>
        bind_cols(species_id = .y)
  ) |>
  left_join(
    sp_info,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot(aes(x = rhat, y = fct_reorder(species_name, rhat))) +
    geom_boxplot() +
    theme_classic() +
    geom_vline(xintercept = 1.01, linetype = 2) +
    ylab('') +
    theme(
      axis.text.y = element_text(face = "italic")
    )
```

# Divergent transitions

```{r divergentTransitions, echo = FALSE,warning=FALSE,message=FALSE}
imap_dfr(
    diag_files,
    ~ readRDS(.x)[['diag_summary']]['num_divergent'] |>
        bind_cols(species_id = .y) |>
        mutate(chain = row_number())
  ) |>
  left_join(
    sp_info |> select(species_id_old, species_name),
    by = c('species_id' = 'species_id_old')
  ) |>
  select(!species_id) |>
  pivot_wider(
    names_from = chain,
    values_from = num_divergent,
    names_prefix = 'chain '
  ) |>
  DT::datatable()
```

# Posterior distribution of parameters

```{r posteriorDist, eval = TRUE,echo = F, fig.height = 7, fig.width = 7,warning=FALSE,message=FALSE}
imap_dfr(
    posteriorPop_files,
    ~ readRDS(.x) |>
      bind_cols(species_id = .y)
  ) |>
  left_join(
    sp_info |> select(species_id_old, species_name),
    by = c('species_id' = 'species_id_old')
  ) ->
post

for(Par in unique(post$par))
{
  print(
    post |>
      filter(par == Par) |>
      ggplot() +
      aes(x = value, y = fct_reorder(species_name, value)) +
      stat_eye() +
      ggtitle(Par) +
      xlab('') +
      theme_classic() +
      ylab('') +
      theme(axis.text.y = element_text(face = "italic"))
  )
}
```

# Correlation between parameters

```{r parCorrelation, eval = TRUE,echo = F, fig.height = 8, fig.width = 8,warning=FALSE,message=FALSE}
parComb <- post |>
  filter(par != 'lp__') |>
  reframe(uqPar = unique(par)) |>
  pull(uqPar) |>
  combn(2)

for(i in 1:ncol(parComb))
{
  print(
    post |>
      filter(par %in% parComb[, i]) |>
      filter(iter %in% sample(iter, 1000)) |>
      select(par, value, iter, species_name) |>
      pivot_wider(names_from = par, values_from = value) |>
      ggplot() +
      aes_string(x = parComb[1, i], y = parComb[2, i]) + 
      geom_point(alpha = .5, size = 0.5) +
      facet_wrap(~ species_name) +
      xlab(parComb[1, i]) +
      ylab(parComb[2, i]) +
      theme_minimal() +
      theme(strip.text = element_text(face = 'italic', size = 7))
  )
}
```

# Predictions

## Mortality probability in function of time

```{r,echo=FALSE,fig.height=6,fig.width=8,warning=FALSE,message=FALSE}
post |>
  pivot_wider(
    names_from = par,
    values_from = value
  ) |>
  group_by(species_name) |>
  expand_grid(deltaYear = seq(0, 1000, 25)) |>
  mutate(
    mort_prob = 1 - (1/(1 + exp(-psi))^deltaYear)
  ) ->
pred_data

pred_data |>
  ggplot() +
  aes(deltaYear, mort_prob) +
  aes(color = species_name) +
  stat_lineribbon(.width = 0.95, alpha = .8) +
  scale_fill_brewer() +
  theme_classic() +
  labs(
    x = 'Delta time (years)',
    y = 'Mortality probability',
    color = ''
  ) +
  geom_text_repel(
    data = pred_data |>
      group_by(species_name) |>
      mutate(
        maxYear = max(deltaYear),
        mort = mean(mort_prob[deltaYear == maxYear])
      ) |>
      slice_head(n = 1),
    mapping = aes(x = maxYear, y = mort, label = species_name),
    max.overlaps = Inf,
    nudge_x = 150, direction = 'y', hjust = 'left',
    fontface = 'italic',
    size = 3
  ) +
  theme(
    legend.position = 'none'
  )
```


## Conspecific competition effect

```{r,echo=FALSE,warning=FALSE,message=FALSE,fig.height=6,fig.width=10}
post |>
  filter(par %in% c('psi', 'Beta')) |>
  pivot_wider(
    names_from = par,
    values_from = value
  ) |>
  group_by(species_name) |>
  expand_grid(BA_comp = seq(0, 100, 5)) |>
  mutate(
    psi_comp = exp(psi + BA_comp * Beta)
  ) ->
pred_data

pred_data |>
  ggplot() +
  aes(BA_comp, psi_comp) +
  aes(color = species_name) +
  stat_lineribbon(.width = 0.9, alpha = .8) +
  scale_fill_brewer() +
  labs(
    x = 'Basal area of larger individuals',
    y = 'Longevity',
    color = ''
  ) +
  theme_classic() +
  geom_text_repel(
    data = pred_data |>
      group_by(species_name) |>
      mutate(
        max_x = max(BA_comp),
        max_y = mean(psi_comp[BA_comp == max_x])
      ) |>
      slice_head(n = 1),
    mapping = aes(x = max_x, y = max_y, label = species_name),
    max.overlaps = Inf,
    nudge_x = 20, direction = 'y', hjust = 'left',
    fontface = 'italic',
    size = 3
  ) +
  theme(
    legend.position = 'none'
  )
```


## Heterospecific competition effect

```{r,echo=FALSE,warning=FALSE,message=FALSE,fig.height=6,fig.width=10}
post |>
  filter(par %in% c('psi', 'Beta', 'theta')) |>
  pivot_wider(
    names_from = par,
    values_from = value
  ) |>
  group_by(species_name) |>
  expand_grid(BA_comp = seq(0, 100, 5)) |>
  mutate(
    psi_comp = exp(psi + (BA_comp * Beta * theta))
  ) ->
pred_data

pred_data |>
  ggplot() +
  aes(BA_comp, psi_comp) +
  aes(color = species_name) +
  stat_lineribbon(.width = 0.9, alpha = .8) +
  scale_fill_brewer() +
  labs(
    x = 'Basal area of larger individuals',
    y = 'Longevity',
    color = ''
  ) +
  theme_classic() +
  geom_text_repel(
    data = pred_data |>
      group_by(species_name) |>
      mutate(
        max_x = max(BA_comp),
        max_y = mean(psi_comp[BA_comp == max_x])
      ) |>
      slice_head(n = 1),
    mapping = aes(x = max_x, y = max_y, label = species_name),
    max.overlaps = Inf,
    nudge_x = 20, direction = 'y', hjust = 'left',
    fontface = 'italic',
    size = 3
  ) +
  theme(
    legend.position = 'none'
  )
```


# Observed vs predicted

```{r crossValidation, echo=FALSE, warning=FALSE, message=FALSE, fig.height=4,fig.width=9}
accur_ls <- mortProp_ls <- mortPropPlot_ls <- R2_ls <- list()

for(Sp in spIds)
{
  val_sp <- dataSource[species_id == Sp]

  # remove plot IDs that do not have random effects
  post_randomEffect <- readRDS(posteriormPlot_files[Sp]) |>
    setDT()

  val_sp <- val_sp[plot_id %in% unique(post_randomEffect$plot_id)]

  # replicate each observation by N_iteration to accommodate the parameters draw
  # Using only 200 iterations of the 8000
  val_sp[, obsID := .I]
  val_sp <- val_sp[rep(obsID, each = 200)]
  sampledIter <- sample(1:(simInfo$maxIter/4 * simInfo$nC), 200)
  val_sp[, iter := sampledIter, by = obsID]

  # merge global parameters
  val_sp[
    post |>
        filter(species_id == Sp) |>
        select(-species_id) |>
        pivot_wider(names_from = par, values_from = value) |>
        as.data.table(),
    `:=`(
      species_name = i.species_name,
      psi = i.psi,
      sigma_plot = i.sigma_plot,
      Beta = i.Beta,
      theta = i.theta
    ),
    on = 'iter'
  ]

  # merge plot specific parameters
  val_sp[
    readRDS(posteriormPlot_files[Sp]) |>
      setDT(),
    psiPlot := i.value,
    on = c('iter', 'plot_id')
  ]

  # random iter were selected, create a new iter vetor to a proprer sequence
  # from 1 to nSample of iteration
  val_sp[, iter2 := 1:.N, by = obsID]

  # compute longevity logit
  val_sp[,
    longev_log := 1/(1 + exp(
        -(
          psi + 
          psiPlot +
          Beta * (BA_comp_sp + theta * BA_comp_inter)
        )
      )
    )
  ]

  # compute mortality rate in function of time
  val_sp[,
    mortality_prob := 1 - (longev_log^deltaYear)
  ]

  # predict mortality event
  val_sp[, mort_random := rbinom(.N, size = 1, mortality_prob)]

# Compute model accuracy
  accur_ls[[Sp]] <- val_sp |>
    group_by(iter2) |>
    reframe(
      TP = sum(mort == 1 & mort_random == 1),
      TN = sum(mort == 0 & mort_random == 0),
      FN = sum(mort == 1 & mort_random == 0),
      FP = sum(mort == 0 & mort_random == 1)
    ) |>
    mutate(
      Acc = (TP + TN)/(TP + TN + FP + FN),
      Sensitivity = TP/(TP + FN),
      Specificity = TN/(FP + TN),
      AccCorrected = (Sensitivity + Specificity)/2
    ) |>
    bind_cols(species_id = Sp)

  # Compute proportion of mortality events (pred vs obs)
  mortProp_ls[[Sp]] <- val_sp |>
    group_by(iter2) |>
    reframe(
      mortProp_obs = sum(mort == 1)/n(),
      mortProp_pred = sum(mort_random == 1)/n()
    ) |>
    bind_cols(species_id = Sp)

  mortPropPlot <- val_sp |>
    group_by(iter2, plot_id) |>
    reframe(
      mortProp_obs = sum(mort == 1)/n(),
      mortProp_pred = sum(mort_random == 1)/n()
    )

  # calculate R2 from regression linear model using the 1:1 obs vs pred
  get_r2 <- function(x, y)
      return(summary(lm(y~x))$r.squared)

  R2_ls[[Sp]] <- as.data.table(mortPropPlot)[,
      .(
        R2 = get_r2(mortProp_obs, mortProp_pred)
      ), 
      by = iter2
    ]

  mortPropPlot_ls[[Sp]] <- mortPropPlot |>
    group_by(plot_id) |>
    reframe(
      mortProp_obs = mean(mortProp_obs),
      mortProp_pred = mean(mortProp_pred)
    ) |>
    bind_cols(species_id = Sp)

  # plot predicted vs observed
  yLim <- c(0, 1)

  val_sp |>
    select(iter2, BA_comp_sp, mort, mort_random) |>
    filter(!is.na(BA_comp_sp)) |>
    mutate(
      varc = cut(
        BA_comp_sp,
        seq(min(BA_comp_sp), max(BA_comp_sp), length.out = 10),
        include.lowest = TRUE
      )
    ) |>
    group_by(varc, iter2) |>
    reframe(
      mort_obs = sum(mort == 1)/n(),
      mort_pred = sum(mort_random == 1)/n()
    ) |>
    pivot_longer(
      cols = contains('mort_')
    ) |>
    mutate(name = gsub('mort_', '', name)) |>
    ggplot() +
    aes(varc, value) +
    aes(color = name) +
    geom_boxplot(alpha = .4) +
    labs(
      x = 'Conspecific basal area',
      y = 'Mortality proportion'
    ) +
    theme_classic() +
    labs(color = '') +
    ylim(yLim) ->
  p1

  val_sp |>
    select(iter2, BA_comp_inter, mort, mort_random) |>
    filter(!is.na(BA_comp_inter)) |>
    mutate(
      varc = cut(
        BA_comp_inter,
        seq(min(BA_comp_inter), max(BA_comp_inter), length.out = 10),
        include.lowest = TRUE
      )
    ) |>
    group_by(varc, iter2) |>
    reframe(
      mort_obs = sum(mort == 1)/n(),
      mort_pred = sum(mort_random == 1)/n()
    ) |>
    pivot_longer(
      cols = contains('mort_')
    ) |>
    mutate(name = gsub('mort_', '', name)) |>
    ggplot() +
    aes(varc, value) +
    aes(color = name) +
    geom_boxplot(alpha = .4) +
    labs(
      x = 'Heterospecific basal area',
      y = 'Mortality proportion'
    ) +
    theme_classic() +
    labs(color = '') +
    ylim(yLim) ->
  p2

  print(
    ggarrange(
      p1, p2,
      nrow = 1,
      ncol = 2,
      labels = unique(val_sp$species_name),
      font.label = list(size = 13, face = 'italic')
    )
  )
}

# save accuracy for model comparison
invisible(
    map_dfr(
    c('accur', 'mortProp', 'mortPropPlot', 'R2'),
    ~ get(paste0(.x, '_ls')) |>
        bind_rows() |>
        saveRDS(file = file.path(
          output_dir,
          paste0(.x, '.RDS')
      )
    )
  )
)
```

## Accuracy

Using four different metrics:

- **Sensitivity** measures the percentage of dead trees that are identified as dead
- **Specificity** measures the percentage of alive trees that are correctly identified as alive
- **Accuracy** is the ratio of number of correct predictions to the total of predictions
- **Accuracy balanced** considers unbalanced accuracy predictions of positive and negative events (sensitivity and specificity)

```{r plot_accur, echo = FALSE, warning=FALSE, message=FALSE, fig.height = 13, fig.width = 9}
sim_toLoad <- c(simInfo$simCompare, simName)

map_dfr(
    sim_toLoad,
    ~ readRDS(file.path('output_sim_processed', simInfo$vitalRates, .x, 'accur.RDS')) |>
      bind_cols(sim = .x)
  ) |>
  left_join(
    sp_info |>
      select(species_id_old, species_name),
    by = c('species_id' = 'species_id_old')
  ) |>
  select(species_name, sim, iter2, Acc, Sensitivity, Specificity, AccCorrected) |>
  pivot_longer(
    cols = c(Acc, Sensitivity, Specificity, AccCorrected)
  ) |>
  mutate(
    name = recode_factor(
      name,
      Acc = 'Accuracy',
      AccCorrected = 'Accuracy corrected',
      Sensitivity = 'Sensitivity',
      Specificity = 'Specificity'
    )
  ) |>
  ggplot() +
  aes(value, fct_reorder(species_name, value)) +
  aes(color = sim) +
  stat_pointinterval(alpha = 0.7) +
  facet_wrap(~ name, scales = 'free_x') +
  theme_classic() +
  labs(
    x = '',
    y = ''
  ) +
  theme(
    legend.position = 'top',
    axis.text.y = element_text(face = 'italic')
  )
```


## Population level mortality rate

```{r plotMortProp,echo=F,warning=FALSE,message=FALSE,fig.height=8,fig.width=7}

map_dfr(
    sim_toLoad,
    ~ readRDS(file.path('output_sim_processed', simInfo$vitalRates, .x, 'mortProp.RDS')) |>
      bind_cols(sim = .x)
  ) |>
  filter(sim == simInfo$simName) |>
  group_by(species_id) |>
  reframe(mprop = unique(mortProp_obs)) |>
  # bind mean of predicted value so we can sort species by the mean of predict
  # this is necessary to match the sorted species for the next figure
  left_join(
    map_dfr(
      sim_toLoad,
      ~ readRDS(file.path('output_sim_processed', simInfo$vitalRates, .x, 'mortProp.RDS')) |>
        bind_cols(sim = .x)
    ) |>
    group_by(species_id) |>
    reframe(mean_pred = mean(mortProp_pred))
  ) |>
  left_join(
    sp_info,
    by = c('species_id' = 'species_id_old')
  ) |>
  arrange(mean_pred) ->
propObs

map_dfr(
    sim_toLoad,
    ~ readRDS(file.path('output_sim_processed', simInfo$vitalRates, .x, 'mortProp.RDS')) |>
      bind_cols(sim = .x)
  ) |>
  left_join(
    sp_info,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(mortProp_pred, y = fct_reorder(species_name, mortProp_pred)) +
  aes(fill = sim) +
  ggridges::geom_density_ridges2(color = NA, alpha = 0.7) +
  geom_segment(
    data = propObs,
    aes(
      x = mprop, xend = mprop,
      y = 1:length(species_name),
      fill = NULL,
      yend = (1:length(species_name)) + .8
    ),
    color = "red", alpha = 0.6
  ) +
  theme_classic() +
  labs(
    x = 'Proportion of mortality events (obs in line vs predicted in dist)',
    y = ''
  ) +
  theme(
    legend.position = 'top',
    axis.text.y = element_text(face = 'italic')
  )
```

```{r plotMortProp2,echo=F,warning=FALSE,message=FALSE,fig.height=10,fig.width=10}
mortPropPlot_ls |>
  bind_rows() |>
  left_join(
    sp_info,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot(aes(x = mortProp_obs, y = mortProp_pred)) +
    geom_point(size = .5, alpha = 0.6) +
    facet_wrap(~species_name) +
    geom_smooth(method = 'gam') +
    geom_abline(slope = 1, intercept = 0) +
    theme_minimal() +
    xlab('Observerd mortality rate') +
    ylab('Predicted mortality rate') +
    theme(
      legend.position = 'top',
      strip.text = element_text(face = 'italic', size = 7)
    )
```

# Leave-one-out cross-validation to compare models

```{r loo,echo = FALSE,warning=FALSE,fig.height = 8, fig.width = 8, warning=FALSE, message=FALSE}
map_dfr(
    spIds,
    ~ map2(
        setNames(sim_toLoad, sim_toLoad),
        .x,
        ~ readRDS(paste0('output_sim_processed/', simInfo$vitalRates, '/', .x, '/loo_', .y, '.RDS'))
      ) |>
      loo::loo_compare() |>
      as.data.frame() |>
      bind_cols(species_id = .x) |>
      rownames_to_column(var = 'sim')
  ) |>
  select(species_id, sim, elpd_diff, se_diff) |>
  pivot_longer(
    cols = c(elpd_diff, se_diff),
    names_to = 'variable',
    values_to = 'value'
  ) |>
  left_join(
    sp_info |>
      select(species_id_old, species_name),
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(x = value, y = fct_reorder(species_name, value), color = sim) +
  geom_point() +
  facet_wrap(~ variable, scales = 'free_x') +
  labs(
    x = '',
    y = '',
    color = 'Model'
  ) +
  theme_classic() +
  theme(legend.position = 'top') +
  theme(axis.text.y = element_text(face = "italic"))
```


```{r loo2,echo = FALSE,warning=FALSE,fig.height = 4, fig.width = 7, warning=FALSE, message=FALSE}
map_dfr(
    spIds,
    ~ map2(
        setNames(sim_toLoad, sim_toLoad),
        .x,
        ~ readRDS(paste0('output_sim_processed/', simInfo$vitalRates, '/', .x, '/loo_', .y, '.RDS'))
      ) |>
      loo::loo_compare() |>
      as.data.frame() |>
      bind_cols(species_id = .x) |>
      rownames_to_column(var = 'sim')
  ) |>
  select(species_id, sim, elpd_diff, se_diff) |>
  pivot_longer(
    cols = c(elpd_diff, se_diff),
    names_to = 'variable',
    values_to = 'value'
  ) |>
  left_join(
    sp_info |>
      select(species_id_old, species_name),
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(value, y = sim) +
  geom_boxplot() +
  facet_wrap(~ variable, scales = 'free_x') +
  theme_classic() +
  ylab('Model')
```
