---
 title: "MCMC diagnostics - recruitment"
 subtitle: "Test v2.0 - Ingrowth model accounting for ingrowth and mortality of new individuals"
 author: "Will Vieira"
 date: "`r paste('Last updated on', format(Sys.time(), '%d %B, %Y'))`"
---

```{r, echo = F, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(cmdstanr))
suppressPackageStartupMessages(library(loo))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(tidybayes))
suppressPackageStartupMessages(library(bayesplot))
suppressPackageStartupMessages(library(ggridges))
suppressPackageStartupMessages(require(RColorBrewer))
suppressPackageStartupMessages(library(DT))
```
```{r load simulations, echo = F}
# Load simulation variables
simInfo <- yaml::read_yaml('_simulation_info.yml')

spIds <- simInfo$spIds
simName <- simInfo$simName
simulations <- simInfo$simulations

# List output files and set vector names to species_id
posteriorPop_files <- grep(
  'posteriorPop',
  dir(file.path('output', simName), full.names = TRUE),
  value = TRUE
)
names(posteriorPop_files) <- names(sort(sapply(spIds, function(x) grep(x, posteriorPop_files))))

diag_files <- grep(
  'diagnostics',
  dir(file.path('output', simName), full.names = TRUE),
  value = TRUE
)
names(diag_files) <- names(sort(sapply(spIds, function(x) grep(x, diag_files))))

# trainData_files <- grep(
#   'trainData',
#   dir(file.path('output', simName), full.names = TRUE),
#   value = TRUE
# )
# names(trainData_files) <- names(sort(sapply(spIds, function(x) grep(x, trainData_files))))
```


```{r load dataset,echo =FALSE}
fec_dt <- readRDS('data/fec_dt.RDS')
fec_dt <- fec_dt[species_id %in% spIds]
```

# Rhat

```{r, echo = F,fig.height = 7, fig.width = 7}
map2_dfr(
  diag_files,
  names(diag_files),
  ~ readRDS(.x)[['rhat']] |>
      bind_cols(species_id = .y)
) |>
ggplot(aes(x = rhat, y = species_id)) +
  geom_boxplot()
```

# Divergent transitions

```{r divergentTransitions, echo = FALSE}
map2_dfr(
  diag_files,
  names(diag_files),
  ~ readRDS(.x)[['diag_summary']]['num_divergent'] |>
      bind_cols(species_id = .y) |>
      mutate(chain = row_number())
) |>
pivot_wider(
  names_from = chain,
  values_from = num_divergent,
  names_prefix = 'chain '
) |>
DT::datatable()
```

# Posterior distribution of parameters

```{r posteriorDist, eval = TRUE,echo = F, fig.height = 7, fig.width = 8}
post <- map2_dfr(
  posteriorPop_files,
  names(posteriorPop_files),
  ~ readRDS(.x) |>
    bind_cols(species_id = .y)
)

for(Par in unique(post$par))
{
  print(
    post |>
      filter(par == Par) |>
      ggplot(aes(x = value, y = species_id)) +
        stat_eye() +
        ggtitle(Par) +
        xlab('')
  )
}
```

# Correlation between parameters

```{r parCorrelation, eval = TRUE,echo = F, fig.height = 8, fig.width = 8}
parComb <- post |>
  summarise(uqPar = unique(par)) |>
  pull(uqPar) |>
  combn(2)

for(i in 1:ncol(parComb))
{
  print(
    post |>
      filter(par %in% parComb[, i]) |>
      filter(iter %in% sample(iter, 1000)) |>
      select(par, value, iter, species_id) |>
      pivot_wider(names_from = par, values_from = value) |>
      ggplot(aes_string(x = parComb[1, i], y = parComb[2, i])) + 
          geom_point() +
          facet_wrap(~ species_id) +
          xlab(parComb[1, i]) +
          ylab(parComb[2, i])
  )
}
```


# Data distribution

```{r organize data, echo = FALSE, warning=FALSE, message=FALSE, fig.height = 7, fig.width = 8}
fec_dt[, .(species_id, nbRecruit, deltaYear_plot)] |>
  pivot_longer(
    cols = c(nbRecruit, deltaYear_plot),
    names_to = 'var'
  ) |>
  ggplot(aes(x = value, y = species_id)) +
    geom_density_ridges(alpha = 0.7, color = NA) +
    facet_wrap(~var, scales = 'free_x') +
    xlab('')
```


# Predictions

```{r,echo=FALSE,fig.height = 6.5, fig.width = 8}
post |>
  pivot_wider(
    names_from = par,
    values_from = value
  ) |>
  group_by(species_id) |>
  expand_grid(deltaYear = seq(0, 50, .5)) |>
  mutate(
    lambda = m * 399 * (1 - p^deltaYear)/(1 - p)
  ) |>
  ggplot(aes(x = deltaYear, y = lambda, color = species_id)) +
        stat_lineribbon(.width = 0.95, alpha = .8) +
        scale_fill_brewer() +
        #theme(legend.position="none") +
        ylab('Number of ingrowth recruits') + xlab('Delta time (years)')
```

# Cross-validation

```{r crossValidation, echo=FALSE, warning=FALSE, message=FALSE, fig.height =4.5, fig.width = 9}
# As we are using each draw of the posterior distribution to predict the validation data, the dataset becames too big to store all species in memory.
# So I will loop over each species to (i) load dataset, (ii) predict for validation, (iii) compute prediction metrics and (iv) generate the plots

# list to save MSE and R2 for each species to be used later
MSE_list <- R2_list <- list()

for(Sp in spIds)
{
  val_sp <- fec_dt[species_id == Sp]

  # remove NAs
  val_sp <- val_sp[!is.na(deltaYear_plot), .(species_id, nbRecruit, deltaYear_plot, plot_size)]

  # replicate each observation by N_iteration to accommodate the parameters draw
  # Using only 1k iterations of the 8000
  val_sp[, obsID := .I]
  val_sp <- val_sp[rep(obsID, each = 1000)]
  sampledIter <- sample(1:(simInfo$maxIter/2 * simInfo$nC), 1000)
  val_sp[, iter := sampledIter, by = obsID]

  # merge global parameters
  val_sp[
    post |>
        filter(species_id == Sp) |>
        select(-species_id) |>
        pivot_wider(names_from = par, values_from = value) |>
        as.data.table(),
    `:=`(
      m = i.m,
      p = i.p
    ),
    on = 'iter'
  ]

  # random iter were selected, create a new iter vetor to a proprer sequence
  # from 1 to nSample of iteration
  val_sp[, iter2 := 1:.N, by = obsID]

  # predict number of recruit
  val_sp[,
    lambda :=  m *
               plot_size *
               (1 - p^deltaYear_plot)/(1 - p)
  ]

  # simulate nbRecruit from model distribution (Poisson dist)
  val_sp[, lambda_random := rpois(.N, lambda)]

  # calculate MSE
  val_sp[, SE := (nbRecruit - lambda_random)^2]
  
  # mean SE over all observations for each iteration and save to list
  MSE_list[[Sp]] <- val_sp[, 
    .(
      MSE = mean(SE, na.rm = TRUE),
      species_id = Sp
    ),
    by = iter2
  ]

  # Calculate R2 (Gelman et al 2018)
  R2gel <- val_sp |>
    group_by(iter2) |>
    summarise(
      var_lambda = var(lambda, na.rm = TRUE),
      mean_lambda = mean(lambda, na.rm = TRUE)      
    ) |>
    bind_cols(
      iter = sampledIter,
      species_id = Sp
    ) |>
    mutate(
      R2gelman = var_lambda/(var_lambda + mean_lambda)
    ) |>
    select(iter, iter2, species_id, R2gelman) |>
    setDT()

  # calculate R2 from regression linear model using the 1:1 obs vs pred
  get_r2 <- function(x, y)
      return(summary(lm(y~x))$r.squared)
  
  R2_list[[Sp]] <- merge(
    R2gel,
    val_sp[,
      .(R2reg = get_r2(nbRecruit, lambda_random)), 
      by = iter2
    ],
    by = 'iter2'
  )

  # plot
  p1 <- val_sp[,.(m_obs=mean(nbRecruit), m_pred=mean(lambda)), by = obsID] |>
    ggplot(aes(m_obs, m_pred)) +
          geom_point(alpha = .4, size = .5) +
          geom_abline(slope = 1, intercept = 0) + 
          tune::coord_obs_pred() +
          theme(
                legend.position="none",
                plot.margin = margin(t = 30)
            ) +
          xlab('Observerd nbRecruit') +
          ylab('Predicted nbRecruit')


  pred_matrix <- val_sp[, .(lambda_random, iter2, obsID)] |>
      pivot_wider(names_from = obsID, values_from = lambda_random) |>
      select(-iter2) |>
      as.matrix()

  xLim_max <- val_sp[, quantile(nbRecruit, probs = 0.99) + 1]
  p2 <- ppc_dens_overlay(
      val_sp[iter2 == 1, nbRecruit],
      pred_matrix[sample(1:nrow(pred_matrix), 200), ]
  ) +
  xlim(0, xLim_max) +
  xlab('nbRecruit')

  print(
    ggpubr::ggarrange(
        p1, p2,
        ncol = 2,
        labels = Sp,
        font.label = list(size = 9, face = 'bold')
    )
  )
}

# save MSE and R2 to be used for model comparison
invisible(
    map_dfr(
    c('MSE', 'R2'),
    ~ get(paste0(.x, '_list')) |>
        bind_rows() |>
        saveRDS(file = file.path(
          'output', 
          simName,
          paste0(.x, '.RDS')
      )
    )
  )
)
```


# Mean squared error (MSE)

```{r plot_MSE, echo = FALSE, warning=FALSE, message=FALSE, fig.height = 7, fig.width = 8}
sim_toLoad <- c(simInfo$simCompare, simName)

map_dfr(
    sim_toLoad,
    ~ readRDS(file.path('output', .x, 'MSE.RDS')) |>
      bind_cols(sim = .x)
  ) |>
  ggplot(aes(x = MSE, y = species_id, fill = sim)) +
    geom_density_ridges(alpha = 0.8, scale= 1.8, color = NA) +
    xlab('Mean Squared error (MSE)') + ylab('')
```


# Rsquared

The distribution of R squared values is calculated using the Gelman et al. [2018](http://www.stat.columbia.edu/~gelman/research/unpublished/bayes_R2.pdf) definition.

```{r plot Rsquared_gelman,echo=FALSE,fig.height = 7.5, fig.width = 8, warning=FALSE, message=FALSE}
map_dfr(
    sim_toLoad,
    ~ readRDS(file.path('output', .x, 'R2.RDS')) |>
      bind_cols(sim = .x)
  ) |>
  ggplot(aes(x = R2gelman, y = species_id, fill = sim)) +
      geom_density_ridges(alpha = 0.8, scale= 1.8, color = NA) +
      xlab('Rsquared (Gelman 2018)')
```

We also calculated the $R^2$ from the linear model using the 1:1 plot comparing observed values to the predicted values.

```{r plot Rsquared_regression,echo=FALSE,fig.height = 6, fig.width = 8, warning=FALSE, message=FALSE}
map_dfr(
    sim_toLoad,
    ~ readRDS(file.path('output', .x, 'R2.RDS')) |>
      bind_cols(sim = .x)
  ) |>
  ggplot(aes(x = R2reg, y = species_id, fill = sim)) +
    geom_density_ridges(alpha = 0.8, scale= 1.8, color = NA) +
    xlab('Rsquared (linear regression)')
```

# Sampling time

```{r plot sampling time, echo = FALSE, fig.height = 7, fig.width = 8}
map2_dfr(
  diag_files,
  names(diag_files),
  ~ readRDS(.x)[['time']][['chains']] |>
    bind_cols(species_id = .y)
) |>
ggplot(aes(x = total/3600, y = species_id)) +
  geom_boxplot() +
  xlab('Fit time (hours)')
```
