---
 title: "MCMC diagnostics - Growth"
 subtitle: "Test v1.0 - Basic growth"
 author: "Will Vieira"
 date: "`r paste('Last updated on', format(Sys.time(), '%d %B, %Y'))`"
---


## Test the convergence of basic functions to the growth rate:

- **sim1** base line growth rate as a function of:
  - discrete canopy status [FALSE/TRUE] as competitive effect
  - gausian function for temperature and precipitation effect
  - lognormal function for size effect

```{r, echo = F}
suppressPackageStartupMessages(library(rstan))
suppressPackageStartupMessages(require(RColorBrewer))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(ggridges))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(DT))
suppressPackageStartupMessages(library(data.table))
```
```{r load simulations, echo = F}
# Load simulation variables
  simInfo <- yaml::read_yaml('_simulation_info.yml')

  # species id
  if(is.null(simInfo$spIds)) {
    spIds <- readRDS('data/spIds.RDS')
  }else {
    spIds <- simInfo$spIds
  }

  simName <- simInfo$simName

  simulations <- simInfo$simulations

  # Human readble simName
  sim1name <- 'sim1'
  #sim2name <- ''
  #sim3name <- ''

#


# Load rStan output and scaleInfo for all species_id
  # get output name for each species_id
  simNames = list.files(path = paste('output', simName, sep = '/'), pattern = '.RDS', recursive = TRUE, full.names = TRUE)
  output = simNames[grep('growthMCMC', simNames)]

  for(sim in simulations) {
    for(sp in spIds) {
      fileName <- output[grep(paste0(#sim,
      sp), output)]
      assign(paste('fit', sim, sp, sep = '_'), readRDS(fileName))
    }
  }

  # get name of parameters for each simulation
  for(sim in simulations)
  {
    # get one MCMC file for a specific sim
    assign(paste0('pars_', sim), rownames(summary(get(ls(pattern = sim)[1]))[[1]]))
  }

#

###########################################
# Diagnostics
###########################################
```



# Rhat

```{r, echo = F, fig.height = 6, fig.width = 10}

  # number of parameters for all simulations
  nbParams <- prod(unlist(lapply(as.list(paste0('pars_', simulations)), function(x) length(get(x)))))
  # create a table with Rhat for each parameter (clumn) and species_id (Rhat)
  Rhat <- data.frame()
  count = 1
  for(sim in simulations) {
    for(sp  in spIds) {
      df <- summary(get(paste('fit', sim, sp, sep = '_')))$summary
      nRow = nrow(df)
      dfToAdd <- data.frame(rep(sim, nRow), rep(sp, nRow), rownames(df), df[, 'Rhat'], row.names = NULL)
      Rhat <- rbind(Rhat ,dfToAdd)
    }
  }
  names(Rhat) <- c('sim', 'sp', 'pars', 'Rhat')

  p <- ggplot(data = Rhat, aes(x = sp, y = Rhat)) +
       geom_boxplot() +
       coord_flip() +
       facet_grid(~sim,
                 labeller = labeller(sim = setNames(c(sim1name),
                                    #sim2name, sim3name),
                                     c('sim1'#,
                                    #sim2', 'sim3'
                                    )))) +
       theme_classic()
  print(p)
```



# Divergent transitions

```{r divergent transitions, echo = FALSE}

divergentDF <- data.frame(matrix(NA, ncol = 3))
names(divergentDF) <- c('sim', 'sp', 'divergent')

count = 1
for(sim in simulations) {
  for(sp in spIds) {
    simNames <- paste('fit', sim, sp, sep = '_')
    sampler_params <- get_sampler_params(get(simNames), inc_warmup = FALSE)
    nbDivergent <- sum(unlist(lapply(sampler_params, function(mat) mat[, 'divergent__'])))
    divergentDF[count, ] <- c(sim, sp, nbDivergent)
    count = count + 1
  }
}
```

```{r divergent table, echo = FALSE}
DT::datatable(divergentDF)
```



# Density plot of each parameter by species id

```{r get MCMC sampling data, echo = F}
  # Prepare data
  # Data frame with sim, sp_id, pars and value
  # Value is all MCMC samples after warmup for all chains together

  mcmcSampleDF <- data.frame()
  for(sim in simulations) {
    for(sp in spIds) {
      simObj <- get(paste('fit', sim, sp, sep = '_'))
      dfToAdd <- gather(as.data.frame(simObj), key = 'pars')
      dfToAdd <- cbind(data.frame(sim = rep(sim, nrow(dfToAdd)), sp = rep(sp, nrow(dfToAdd))), dfToAdd)
      mcmcSampleDF <- rbind(mcmcSampleDF, dfToAdd)
    }
  }
```

```{r plot mcmcSample by pars, echo = F, message=FALSE, fig.height = 7, fig.width = 9}

  for(par in unique(mcmcSampleDF$pars))
  {
    p <- ggplot(data = subset(mcmcSampleDF, pars == par), aes(value, sp)) +
         geom_density_ridges_gradient() +
         facet_grid(~sim, scales = ifelse(par %in% c("T_opt", "sigmaT_opt", "P_opt", "sigmaP_opt"), 'free_x', 'fixed'),
                   labeller = labeller(sim = setNames(c(sim1name#,
                                       #sim2name, sim3name
                                       ),
                                       c('sim1'#, 'sim2', 'sim3'
                                       )))) +
         theme_classic() +
         xlab(par)
    print(p)
  }
```

# Data distribution

```{r organize data, echo = FALSE}
growth_dt <- readRDS('data/growth_dt.RDS')
growth_dt <- growth_dt[species_id %in% spIds]
growth_dt <- growth_dt[, .(species_id, growth, dbh0, mean_temp_period_3_lag, tot_pp_period3_lag, canopyStatus)]

# transform dt
growth <- tidyr::gather(data = growth_dt,
            growth, dbh0, mean_temp_period_3_lag, tot_pp_period3_lag, canopyStatus,
            key = 'var', value = 'value')
rm(growth_dt)
```

```{r plot data distribution, echo = FALSE, message=FALSE, fig.height = 7, fig.width = 10}
p <- ggplot(data = growth, aes(value, species_id)) +
      geom_density_ridges_gradient() +
      facet_grid(~var, scales = 'free_x') + #,
      #          labeller = labeller(sim = setNames(c(sim1name,
      #                              sim2name),
      #                              c('sim1', 'sim2')))) +
      theme_classic()
print(p)
```


# Size, temperature, precipitation and competition effect

```{r, echo = F, fig.height = 11, fig.width = 9}
# Plot of size, temperature, precipitation and competition effect

  # define functions of the model
    # size effect
    sizeEffect = function(x, Phi_opt, sigmaPhi_opt) {
      lgPart <- (log(x/Phi_opt)/sigmaPhi_opt)^2
      return( exp(-lgPart))
    }

    # temperature effect
    tempEffect = function(xtemp, T_opt, sigmaT_opt) {
      return(0.0001 + exp(-0.5*(xtemp - T_opt)^2/sigmaT_opt^2))
    }

    # precipitation effect
    precEffect = function(xprec, P_opt, sigmaP_opt) {
      return(exp(-0.5*(xprec - P_opt)^2/sigmaP_opt^2))
    }

    # competition effect
    # compEffect = function(xcanopy, sim, param) {

    #   if(sim == 'sim1') {
    #     Lo = 0
    #     Beta = 0.5
    #     Mid = param['Mid', 1]
    #   }else if(sim == 'sim2') {
    #     Lo = param['Lo', 1]
    #     Beta = 0.5
    #     Mid = param['Mid', 1]
    #   }else {
    #     Lo = param['Lo', 1]
    #     Beta = param['Beta', 1]
    #     Mid = param['Mid', 1]
    #   }

    #   return( (Lo + ((1 - Lo) / (1 + exp(-Beta * (xcanopy - Mid))))) )
    # }

  #

  ## PLOT ##

  # define x axis
  sizex = seq(0, 700, 0.5)
  tempx = seq(0, 20)
  precx = seq(0, 800)
  #canopyDist = seq(-35, 40, length.out = 200)

  # species specfic colours
  spCols <- rev(RColorBrewer::brewer.pal(length(spIds), "Set1"))

  par(mfrow = c(4, length(simulations)), xaxs = "i", las = 1, mar = c(3, 1, 3, 1.5), oma = c(1,3.5,2,.2), mgp = c(1.5, 0.3, 0), tck = -.008, cex.lab = 1.2, cex.axis = 1.2)

  # plot size effect
  for(sim in simulations) {
    for(sp in 1:length(spIds)) {
      # get parameters
      out <- summary(get(paste('fit', sim, spIds[sp], sep = '_')))$summary
      # plot
      if(sp == 1) plot(sizex, sizeEffect(sizex, out['Phi_opt', 1], out['sigmaPhi_opt', 1]),
                  type = 'l', ylim = c(0, 1), xlab = 'Size (mm)', ylab = 'Size effect',
                  col = spCols[sp],
                  yaxt = ifelse(sim == 'sim1', "s", "n"))
      if(sim == 'sim2') axis(2, labels = FALSE)
      points(sizex, sizeEffect(sizex, out['Phi_opt', 1], out['sigmaPhi_opt', 1]), type = 'l', col = spCols[sp])
    }
    mtext(ifelse(sim == 'sim1', sim1name, ifelse(sim == 'sim2', sim2name, sim3name)), side = 3, line = 0)
    if(sim == 'sim1') mtext("Size effect", 2, las = 0, cex = .8, line = 1.9)
  }

  # plot temp effect
  for(sim in simulations) {
    for(sp in 1:length(spIds)) {
      # get parameters
      out <- summary(get(paste('fit', sim, spIds[sp], sep = '_')))$summary
      topt <- out['T_opt', 1]
      sigTopt <- out['sigmaT_opt', 1]

      if(sp == 1) plot(tempx, tempEffect(tempx, topt, sigTopt), type = 'l',
                  ylim = c(0, 1), xlab = 'Temperature (°C)', ylab = 'Temperature effect',
                  col = spCols[sp],
                  yaxt = ifelse(sim == 'sim1', "s", "n"))
      if(sim == 'sim2') axis(2, labels = FALSE)
      points(tempx, tempEffect(tempx, topt, sigTopt), type = 'l', col = spCols[sp])
    }
    if(sim == 'sim1') mtext("Temperature effect", 2, las = 0, cex = .8, line = 1.9)
  }

  # plot precipitation effect
  for(sim in simulations) {
    for(sp in 1:length(spIds)) {
      # get parameters
      out <- summary(get(paste('fit', sim, spIds[sp], sep = '_')))$summary
      popt <- out['P_opt', 1]
      sigPopt <- out['sigmaP_opt', 1]

      if(sp == 1) plot(precx, precEffect(precx, popt, sigPopt), type = 'l',
                  ylim = c(0, 1), xlab = 'Precipiation (mm)', ylab = 'Precipiation effect',
                  col = spCols[sp],
                  yaxt = ifelse(sim == 'sim1', "s", "n"))
      if(sim == 'sim2') axis(2, labels = FALSE)
      points(precx, precEffect(precx, popt, sigPopt), type = 'l', col = spCols[sp])
    }
    if(sim == 'sim1') mtext("Precipiation effect", 2, las = 0, cex = .8, line = 1.9)
  }

  # plot canopy effect
  # for(sim in simulations) {
  #   for(sp in 1:length(spIds)) {
  #     # get parameters
  #     out <- summary(get(paste('fit', sim, spIds[sp], sep = '_')))$summary

  #     if(sp == 1) plot(canopyDist, compEffect(canopyDist, sim, param = out), xlim = c(-35, 40),
  #     ylim = c(0, 1), type = 'l', xlab = 'Canopy distance', ylab = 'Competition effect',
  #     col = spCols[sp],
  #     yaxt = ifelse(sim == 'sim1', "s", "n"))
  #     if(sim == 'sim2') axis(2, labels = FALSE)

  #     points(canopyDist, compEffect(canopyDist, sim, param = out), type = 'l', col = spCols[sp])
  #   }

  #   if(sim == 'sim1') mtext("Competition effect", 2, las = 0, cex = .8, line = 1.9)
  # }
  plot(0, xlim = c(-.5, 3), ylim = c(0, 1), pch = '', xlab = 'canopy status', ylab = 'Canopy effect', xaxt = 'n')
  axis(1, labels = T, at = c(0, 1))
  for(sp in 1:length(spIds))
  {
    points(0, summary(get(paste('fit', sim, spIds[sp], sep = '_')))$summary['beta', 1], pch = 8, col = spCols[sp])
  }
  legend('bottomright', legend = spIds, col = spCols, lty = 1, bty = 'n', cex = 0.9)
```


# Predictions

```{r, echo = F}
# Plot predictions
  # define model

  growth_model <- function(param, sim, newdata) {

    # # pars depending on simulation
    # if(sim == 'sim1') {
    #   Lo = 0
    #   Beta = 0.5
    #   Mid = param['Mid']
    # }else if(sim == 'sim2') {
    #   Lo = param['Lo']
    #   Beta = 0.5
    #   Mid = param['Mid']
    # }else {
    #   Lo = param['Lo']
    #   Beta = param['Beta']
    #   Mid = param['Mid']
    # }

    lgPart <- (log(newdata$D/param["Phi_opt"])/param["sigmaPhi_opt"])^2

    mu_d <- param["pdg"] *
            (newdata$CS + (1 - newdata$CS) * param["beta"]) *
            exp(-0.5*(newdata$TP - param["T_opt"])^2/param["sigmaT_opt"]^2) *
            exp(-0.5*(newdata$PP - param["P_opt"])^2/param["sigmaP_opt"]^2) *
            exp(-lgPart)
    
    growth <- mu_d
    
    growth_dt <- cbind(growth, newdata)
  }

  # data frame with fixed data to run the model
  # Temp
  newdataT <- data.frame(expand.grid(CS = 1,
                                     TP = seq(-30, 20, length.out = 300),
                                     PP = 1150,
                                     D = c(100, 200, 400)))
  # Prec
  newdataP <- data.frame(expand.grid(CS = 1,
                                     TP = 1.5,
                                     PP = seq(200, 2500, length.out = 300),
                                     D = c(100, 200, 400)))
  # Size
  newdataD <- data.frame(expand.grid(CS = 1,
                                     TP = 1.5,
                                     PP = 1150,
                                     D = seq(10, 900, length.out = 300)))

  # Canopy distance
  newdataC <- data.frame(expand.grid(CS = seq(-35, 40, length.out = 300),
                                     TP = 1.5,
                                     PP = 1150,
                                     D = c(100, 200, 400)))

  # Function in which you give a rstan output (species_id) and it returns the three plots
  plot_prediction <- function(sim, sp, model, data) {

    # Prepare data
    newdataT <- data[['newdataT']]
    newdataP <- data[['newdataP']]
    newdataD <- data[['newdataD']]
    newdataC <- data[['newdataC']]

    # simulation
    fit <- get(paste('fit', sim, sp, sep = '_'))

    # posterior distribution
    posterior <- rstan::extract(fit)

    param_summary <- summary(fit, probs = c(0.1, 0.9))$summary
    rownames(param_summary) <- names(posterior)
    param_summary <- param_summary[,-c(2:3)]

    # define parameters
    nRow <- nrow(param_summary)
    param_mean <- param_summary[-nRow, 1]
    param_ci10 <- param_summary[-nRow, 2]
    param_ci90 <- param_summary[-nRow, 3]

    # get predictions
    growth_meanT <- growth_model(param = param_mean, sim = sim, newdata = newdataT)
    growth_ci10T <- growth_model(param = param_ci10, sim = sim, newdata = newdataT)
    growth_ci90T <- growth_model(param = param_ci90, sim = sim, newdata = newdataT)

    growth_meanP <- growth_model(param = param_mean, sim = sim, newdata = newdataP)
    growth_ci10P <- growth_model(param = param_ci10, sim = sim, newdata = newdataP)
    growth_ci90P <- growth_model(param = param_ci90, sim = sim, newdata = newdataP)

    growth_meanD <- growth_model(param = param_mean, sim = sim, newdata = newdataD)
    growth_ci10D <- growth_model(param = param_ci10, sim = sim, newdata = newdataD)
    growth_ci90D <- growth_model(param = param_ci90, sim = sim, newdata = newdataD)

    growth_meanC <- growth_model(param = param_mean, sim = sim, newdata = newdataC)
    growth_ci10C <- growth_model(param = param_ci10, sim = sim, newdata = newdataC)
    growth_ci90C <- growth_model(param = param_ci90, sim = sim, newdata = newdataC)


    size2plot <- unique(newdataT$D)
    colsize <- rev(RColorBrewer::brewer.pal(length(size2plot) + 1, "PuBuGn"))

    if(sim == 'sim1')
      par(mfcol = c(4, length(simulations)), xaxs="i", las = 1, mar = c(3, 1, 3, 1.5), oma = c(1,3.5,2,.2), mgp = c(1.5, 0.3, 0), tck = -.008, cex.lab = 1.2, cex.axis = 1.2)

    # Growth vs temperature
    plot(growth ~ TP, growth_ci10T, type = "n", ylim = c(0, 1),
         xlab = "Temperature (°C)", ylab = "", cex.lab = 1.2, yaxt = ifelse(sim == 'sim1', "s", "n"))
    if(sim == 'sim2') axis(2,labels=F)

    for(dhp in size2plot) {
      mu <- subset(growth_meanT, D == dhp)
      ci10 <- subset(growth_ci10T, D == dhp)
      ci90 <- subset(growth_ci90T, D == dhp)
      mycol <- colsize[size2plot == dhp]

      lines(growth ~ TP, mu, cex=.5, pch=19, col = mycol)
      polygon(c(mu$TP, rev(mu$TP)),
              c(ci10$growth, rev(ci90$growth)),
              col = alpha(mycol, 0.5), border = NA)
    }
    if(sim == 'sim1')
      legend("topleft", legend = size2plot, title = "Size",
             fill = colsize[1:3], bty = "n", border = "transparent")

     mtext(ifelse(sim == 'sim1', sim1name, ifelse(sim == 'sim2', sim2name, sim3name)), side = 3, line = 0)
     mtext(sp, side = 3, outer = TRUE, cex = 1, line = -1.8)

    # Growth vs precipitation
    plot(growth ~ PP, growth_ci10P, type = "n", ylim = c(0, 1),
         xlab = "Precipiation (mm)", ylab = " ", yaxt = ifelse(sim == 'sim1', "s", "n"), cex.lab = 1.2)
    axis(2, labels = F)
    for(dhp in size2plot) {
      mu <- subset(growth_meanP, D == dhp)
      ci10 <- subset(growth_ci10P, D == dhp)
      ci90 <- subset(growth_ci90P, D == dhp)
      mycol <- colsize[size2plot == dhp]

      lines(growth ~ PP, mu, cex = .5, pch = 19, col = mycol)
      polygon(c(mu$PP, rev(mu$PP)),
              c(ci10$growth, rev(ci90$growth)),
              col = alpha(mycol, 0.5), border = NA)
    }
    mtext("P(Growth)", 2, outer = T, las = 0, cex = .8, line = 1.5)

    # Growth vs size
    plot(growth ~ D, growth_ci10D, type = "n", ylim = c(0, 1),
         xlab = "Size (mm)", ylab = "", yaxt = ifelse(sim == 'sim1', "s", "n"), cex.lab = 1.2)
    axis(2, labels = F)
    lines(growth ~ D, growth_meanD, cex=.5, pch=19, col = "grey25")
    polygon(c(growth_meanD$D, rev(growth_meanD$D)),
              c(growth_ci10D$growth, rev(growth_ci90D$growth)),
              col = alpha("grey25", 0.1), border = NA)

    # Growth vs canopy
    plot(growth ~ CS, growth_ci10C, type = "n", ylim = c(0, 1),
         xlab = "Canopy distance from S* (m)", ylab = "", yaxt = ifelse(sim == 'sim1', "s", "n"), cex.lab = 1.2)
    axis(2, labels = F)
    for(dhp in size2plot) {
      mu <- subset(growth_meanC, D == dhp)
      ci10 <- subset(growth_ci10C, D == dhp)
      ci90 <- subset(growth_ci90C, D == dhp)
      mycol <- colsize[size2plot == dhp]

      lines(growth ~ CS, mu, cex = .5, pch = 19, col = mycol)
      polygon(c(mu$CS, rev(mu$CS)),
              c(ci10$growth, rev(ci90$growth)),
              col = alpha(mycol, 0.5), border = NA)
    }
  }
  ```

  ```{r plot predictions,echo=FALSE,fig.height = 12, fig.width = 9}
  # plot all species
  for(sp in spIds) {
    for(sim in simulations) {
      plot_prediction(sim, sp, growth_model,
      data = list(newdataT = newdataT, newdataP = newdataP, newdataD = newdataD, newdataC = newdataC))
    }
  }
  #
#
```



# Sampling time

```{r sampling time, echo = FALSE}
samplingDF <- data.frame()

for(sim in simulations) {
  for(sp in spIds) {
    simNames <- paste('fit', sim, sp, sep = '_')
    sumTime <- rowSums(get_elapsed_time(get(simNames)))
    nChain <- length(sumTime)
    df1 <- data.frame(sim = rep(sim, nChain),
                     sp = rep(sp, nChain),
                     time = sumTime)
    samplingDF <- rbind(samplingDF, df1)
  }
}

samplingDF$time <- samplingDF$time/60/60
```

```{r plot sampling time, echo = FALSE, fig.height = 6, fig.width = 12}

p <- ggplot(data = samplingDF, aes(x = sim, y = time)) +
      geom_boxplot() +
      facet_grid(~sp) +
      theme_classic() +
      ylab('Time (h)')
print(p)
```
```{r plot sampling time2, echo = FALSE, fig.height = 5, fig.width = 6}
p <- ggplot(data = samplingDF, aes(x = sim, y = time)) +
      geom_boxplot() +
      theme_classic() +
      ylab('Time (h)')
print(p)
```
