---
 title: "MCMC diagnostics - growth"
 subtitle: "Model: intercept + plot random effects"
 author: "Will Vieira"
 date: "`r paste('Last updated on', format(Sys.time(), '%d %B, %Y'))`"
---

```{r, echo = F, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(cmdstanr))
suppressPackageStartupMessages(library(loo))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(tidybayes))
suppressPackageStartupMessages(library(bayesplot))
suppressPackageStartupMessages(library(ggridges))
suppressPackageStartupMessages(require(RColorBrewer))
suppressPackageStartupMessages(library(DT))
suppressPackageStartupMessages(library(ggrepel))
```
```{r load simulations, echo = F,warning=FALSE,message=FALSE}
# Load simulation variables
model_dir <- 'model_growth/intcpt_plot'
simInfo <- yaml::read_yaml(file.path(model_dir, '_simulation_info.yml'))
output_dir <- file.path('output_sim_processed', simInfo$vitalRates, simInfo$simName)

sp_info <- read_csv(file.path('..', 'forest-IPM', 'data', 'species_id.csv'))
spIds <- simInfo$spIds
simName <- simInfo$simName

# List output files and set vector names to species_id
posteriorPop_files <- grep(
  'posterior_pop',
  dir(output_dir, full.names = TRUE),
  value = TRUE
)
names(posteriorPop_files) <- names(sort(sapply(spIds, function(x) grep(x, posteriorPop_files))))

posteriormPlot_files <- grep(
  'posterior_plot',
  dir(output_dir, full.names = TRUE),
  value = TRUE
)
names(posteriormPlot_files) <- names(sort(sapply(spIds, function(x) grep(x, posteriormPlot_files))))

diag_files <- grep(
  'diagnostics',
  dir(output_dir, full.names = TRUE),
  value = TRUE
)
names(diag_files) <- names(sort(sapply(spIds, function(x) grep(x, diag_files))))
```


```{r load dataset,echo =FALSE,warning=FALSE,message=FALSE}
dataSource <- readRDS(
  gsub('../../', '', simInfo$dataSources)
)
```

# Rhat

```{r, echo = F,fig.height = 7, fig.width = 7,warning=FALSE,message=FALSE}
imap_dfr(
    diag_files,
    ~ readRDS(.x)[['rhat']] |>
        bind_cols(species_id = .y)
  ) |>
  left_join(
    sp_info,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot(aes(x = rhat, y = fct_reorder(species_name, rhat))) +
    geom_boxplot() +
    theme_classic() +
    geom_vline(xintercept = 1.01, linetype = 2) +
    ylab('') +
    theme(
      axis.text.y = element_text(face = "italic")
    )
```

# Divergent transitions

```{r divergentTransitions, echo = FALSE,warning=FALSE,message=FALSE}
imap_dfr(
    diag_files,
    ~ readRDS(.x)[['diag_summary']]['num_divergent'] |>
        bind_cols(species_id = .y) |>
        mutate(chain = row_number())
  ) |>
  left_join(
    sp_info |> select(species_id_old, species_name),
    by = c('species_id' = 'species_id_old')
  ) |>
  select(!species_id) |>
  pivot_wider(
    names_from = chain,
    values_from = num_divergent,
    names_prefix = 'chain '
  ) |>
  DT::datatable()
```

# Posterior distribution of parameters

```{r posteriorDist, eval = TRUE,echo = F, fig.height = 7, fig.width = 7,warning=FALSE,message=FALSE}
imap_dfr(
    posteriorPop_files,
    ~ readRDS(.x) |>
      bind_cols(species_id = .y)
  ) |>
  left_join(
    sp_info |> select(species_id_old, species_name),
    by = c('species_id' = 'species_id_old')
  ) ->
post

for(Par in unique(post$par))
{
  print(
    post |>
      filter(par == Par) |>
      ggplot() +
      aes(x = value, y = fct_reorder(species_name, value)) +
      stat_eye() +
      ggtitle(Par) +
      xlab('') +
      theme_classic() +
      ylab('') +
      theme(axis.text.y = element_text(face = "italic"))
  )
}
```

# Correlation between parameters

```{r parCorrelation, eval = TRUE,echo = F, fig.height = 8, fig.width = 8,warning=FALSE,message=FALSE}
parComb <- post |>
  filter(par != 'lp__') |>
  reframe(uqPar = unique(par)) |>
  pull(uqPar) |>
  combn(2)

for(i in 1:ncol(parComb))
{
  print(
    post |>
      filter(par %in% parComb[, i]) |>
      filter(iter %in% sample(iter, 1000)) |>
      select(par, value, iter, species_name) |>
      pivot_wider(names_from = par, values_from = value) |>
      ggplot() +
      aes_string(x = parComb[1, i], y = parComb[2, i]) + 
      geom_point(alpha = .5, size = 0.5) +
      facet_wrap(~ species_name) +
      xlab(parComb[1, i]) +
      ylab(parComb[2, i]) +
      theme_minimal() +
      theme(strip.text = element_text(face = 'italic', size = 7))
  )
}
```

# Predictions

## Size in function of time

```{r,echo=FALSE,fig.height=6,fig.width=8,warning=FALSE,message=FALSE}
post |>
  filter(par %in% c('r', 'Lmax')) |>
  pivot_wider(
    names_from = par,
    values_from = value
  ) |>
  group_by(species_name) |>
  expand_grid(time = seq(0, 1000, 50)) |>
  mutate(
    Lt = Lmax * (1 - exp(-exp(r) * time))
  ) ->
pred_data

pred_data |>
  ggplot() +
  aes(time, Lt) +
  aes(color = species_name) +
  stat_lineribbon(.width = 0.95, alpha = .8) +
  scale_fill_brewer() +
  theme_classic() +
  labs(
    x = 'Time (years)',
    y = 'DBH size (mmm)',
    color = ''
  ) +
  geom_text_repel(
    data = pred_data |>
      group_by(species_name) |>
      mutate(
        max_x = max(time),
        max_y = mean(Lt[time == max_x])
      ) |>
      slice_head(n = 1),
    mapping = aes(x = max_x, y = max_y, label = species_name),
    max.overlaps = Inf,
    nudge_x = 200, direction = 'y', hjust = 'left',
    fontface = 'italic',
    size = 3
  ) +
  theme(
    legend.position = 'none'
  )
```


# Observed vs predicted

```{r crossValidation, echo=FALSE, warning=FALSE, message=FALSE, fig.height=4,fig.width=9}
# list to save MSE and R2 for each species to be used later
MSE_list <- R2_list <- list()

for(Sp in spIds)
{
  val_sp <- dataSource[species_id == Sp]

  ## compute de deltaTime between measures of dbh
  val_sp[,
    deltaTime := year_measured - lag(year_measured, 1),
    by = tree_id
  ]

  ## define previous measure
  val_sp[,
    dbh0 := lag(dbh, 1),
    by = tree_id
  ]

  ## Fill the NA first measures with their non lag information
  val_sp[is.na(deltaTime), deltaTime := 0]
  val_sp[deltaTime == 0, dbh0 := dbh]

  # remove NAs
  val_sp <- na.omit(
    val_sp[,
      .(
        plot_id, species_id, deltaTime, dbh, dbh0, BA_comp_sp,
        BA_comp_inter, bio_01_mean_scl, bio_12_mean_scl
      )
    ]
  )

  # replicate each observation by N_iteration to accommodate the parameters draw
  # Using only 200 iterations of the 8000
  val_sp[, obsID := .I]
  val_sp <- val_sp[rep(obsID, each = 200)]
  sampledIter <- sample(1:(simInfo$maxIter/4 * simInfo$nC), 200)
  val_sp[, iter := sampledIter, by = obsID]

  # merge global parameters
  val_sp[
    post |>
      filter(species_id == Sp) |>
      select(-species_id) |>
      pivot_wider(names_from = par, values_from = value) |>
      as.data.table(),
    `:=`(
      r = i.r,
      Lmax = i.Lmax,
      sigma_obs = i.sigma_obs,
      sigma_plot = i.sigma_plot
      # Beta = i.Beta,
      # theta = i.theta,
      # optimal_temp = i.optimal_temp,
      # tau_temp = i.tau_temp,
      # optimal_prec = i.optimal_prec,
      # tau_prec = i.tau_prec
    ),
    on = 'iter'
  ]

  # merge plot specific parameters
  val_sp[
    readRDS(posteriormPlot_files[Sp]) |>
      setDT(),
    rPlot_log := i.value,
    on = c('iter', 'plot_id')
  ]

  # remove NA plot_id_seq
  val_sp <- val_sp[!is.na(rPlot_log)]

  # random iter were selected, create a new iter vetor to a proprer sequence
  # from 1 to nSample of iteration
  val_sp[, iter2 := 1:.N, by = obsID]

  # compute r plot specific parameter
  val_sp[,
    r_plot := exp(
      r +
      rPlot_log
      # Beta * (BA_comp_sp + theta * BA_comp_intra) +
      # -tau_temp * (bio_01_mean_scl - optimal_temp)^2 +
      # -tau_prec * (bio_12_mean_scl - optimal_prec)^2
    )
  ]

  # predict size
  val_sp[,
    pred_size := dbh0 *
                 exp(-r_plot * deltaTime) +
                 Lmax * (1 -exp(-r_plot * deltaTime))
  ]

  # simulate size from model distribution (normal dist)
  val_sp[, pred_size_random := rnorm(.N, pred_size, sigma_obs)]

  # calculate observed and predicted growth 
  val_sp[, obs_growth := (dbh - dbh0)/deltaTime]
  val_sp[, pred_growth := (pred_size - dbh0)/deltaTime]
  val_sp[, pred_growth_random := (pred_size_random - dbh0)/deltaTime]

  # calculate MSE
  val_sp[, SE_size := (dbh - pred_size_random)^2]
  val_sp[, SE_growth := (obs_growth - pred_growth_random)^2]

  # mean SE over all observations for each iteration and save to list
  MSE_list[[Sp]] <- val_sp[, 
    .(
      MSE_size = mean(SE_size, na.rm = TRUE),
      MSE_growth = mean(SE_growth, na.rm = TRUE),
      species_id = Sp
    ),
    by = iter2
  ]

  # Calculate R2 (Gelman et al 2018)
  R2gel <- val_sp |>
    group_by(iter2) |>
    summarise(var_size = var(pred_size, na.rm = TRUE)) |>
    bind_cols(iter = sampledIter) |>
    left_join(
      post |>
        filter(species_id == Sp & par == 'sigma_obs') |>
        mutate(sigma_obs = value) |>
        select(iter, sigma_obs, species_id)
    ) |>
    mutate(R2gelman_size = var_size/(var_size + sigma_obs^2)) |>
    select(iter, iter2, species_id, R2gelman_size) |>
    setDT()

  # calculate R2 from regression linear model using the 1:1 obs vs pred
  get_r2 <- function(x, y)
      return(summary(lm(y~x))$r.squared)
  
  R2_list[[Sp]] <- merge(
    R2gel,
    val_sp[,
      .(
        R2reg_size = get_r2(dbh, pred_size_random),
        R2reg_growth = get_r2(obs_growth, pred_growth_random),
        R2reg_size_mean = get_r2(dbh, pred_size),
        R2reg_growth_mean = get_r2(obs_growth, pred_growth)
      ), 
      by = iter2
    ],
    by = 'iter2'
  )

  # define species specific axis limits for growth
  yLim <- val_sp[, quantile(obs_growth, probs = c(0.001, .999), na.rm = T)]
  
  # plot
  p1 <- val_sp[
      deltaTime > 0,
      .(m_obs=mean(dbh), m_pred=mean(pred_size_random)),
      by = obsID
    ] |>
    ggplot() +
    aes(m_obs, m_pred) +
    geom_point(alpha = .3, size = .5) +
    geom_abline(slope = 1, intercept = 0) + 
    geom_smooth(method = 'gam') +
    tune::coord_obs_pred() +
    theme_classic() +
    theme(
      legend.position = 'none',
      plot.margin = margin(t = 30)
    ) +
    xlab('Observerd size (mm)') +
    ylab('Predicted size (mm)')

  pred_matrix <- val_sp[
      deltaTime > 0,
      .(pred_size_random, iter2, obsID)
    ] |>
    pivot_wider(names_from = obsID, values_from = pred_size_random) |>
    select(-iter2) |>
    as.matrix()

  p2 <- val_sp[
      deltaTime > 0,
      .(m_obs = mean(obs_growth), m_pred = mean(pred_growth)),
      by = obsID
    ] |>
    ggplot() +
    aes(m_obs, m_pred) +
    geom_hex(bins = 100) +
    geom_abline(slope = 1, intercept = 0) +
    xlim(yLim) + ylim(yLim) +
    theme_minimal() +
    theme(
      legend.position="none",
      plot.margin = margin(t = 30)
    ) +
    xlab('Observerd growth') +
    ylab('Predicted growth')

  pred_size_matrix <- val_sp[deltaTime > 0 & !is.na(pred_size_random), .(pred_size_random, iter2, obsID)] |>
      pivot_wider(names_from = obsID, values_from = pred_size_random) |>
      select(-iter2) |>
      as.matrix()

  p3 <- ppc_dens_overlay(
      val_sp[iter2 == 1 & deltaTime > 0, dbh],
      pred_size_matrix
  ) +
  xlab('Obs vs predicted size') +
  ylab('') +
  theme_classic()

  pred_growth_matrix <- val_sp[deltaTime > 0 & !is.na(obs_growth) & !is.na(pred_growth), .(pred_growth_random, iter2, obsID)] |>
      pivot_wider(names_from = obsID, values_from = pred_growth_random) |>
      select(-iter2) |>
      as.matrix()

  p4 <- ppc_dens_overlay(
      val_sp[iter2 == 1 & deltaTime > 0 & !is.na(obs_growth), obs_growth],
      pred_growth_matrix
  ) +
  xlab('Obs vs predicted growth') +
  xlim(yLim) +
  ylab('') +
  theme_classic()

  print(
    ggpubr::ggarrange(
        p1, p2, p3, p4,
        ncol = 2, nrow = 2,
        labels = unique(val_sp$species_name),
        font.label = list(size = 13, face = 'italic')
    )
  )
}

# save MSE and R2 to be used for model comparison
invisible(
    map_dfr(
    c('MSE', 'R2'),
    ~ get(paste0(.x, '_list')) |>
        bind_rows() |>
        saveRDS(file = file.path(
          output_dir,
          paste0(.x, '.RDS')
      )
    )
  )
)
```


# Mean squared error (MSE)

```{r plot_MSE, echo = FALSE, warning=FALSE, message=FALSE, fig.height = 7, fig.width = 8}
sim_toLoad <- c(simInfo$simCompare, simName)

map_dfr(
    sim_toLoad,
    ~ readRDS(file.path('output_sim_processed', simInfo$vitalRates, .x, 'MSE.RDS')) |>
      bind_cols(sim = .x)
  ) |>
  left_join(
    sp_info |>
      select(species_id_old, species_name),
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(x = MSE, y = fct_reorder(species_name, MSE), fill = sim) +
  geom_density_ridges(alpha = 0.8, scale= 1.8, color = NA) +
  labs(
    x = 'Mean Squared error (MSE)',
    y = '',
    fill = 'Model'
  ) +
  theme_classic() +
  theme(axis.text.y = element_text(face = "italic"))
```


# Rsquared

The distribution of R squared values is calculated using the Gelman et al. [2018](http://www.stat.columbia.edu/~gelman/research/unpublished/bayes_R2.pdf) definition.

```{r plot Rsquared_gelman,echo=FALSE,fig.height = 7, fig.width = 7, warning=FALSE, message=FALSE}
map_dfr(
    sim_toLoad,
    ~ readRDS(file.path('output_sim_processed', simInfo$vitalRates, .x, 'R2.RDS')) |>
      bind_cols(sim = .x)
  ) |>
  left_join(
    sp_info |>
      select(species_id_old, species_name),
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(x = R2gelman, y = fct_reorder(species_name, R2gelman), fill = sim) +
  geom_density_ridges(alpha = 0.8, scale= 1.8, color = NA) +
  labs(
    x = 'Rsquared (Gelman 2018)',
    y = '',
    fill = 'Model'
  ) +
  theme_classic() +
  theme(axis.text.y = element_text(face = "italic"))
```

We also calculated the pseudo $R^2$ from the linear model using the 1:1 plot comparing observed values to the predicted values.

```{r plot Rsquared_regression,echo=FALSE,fig.height = 7, fig.width = 7, warning=FALSE, message=FALSE}
map_dfr(
    sim_toLoad,
    ~ readRDS(file.path('output_sim_processed', simInfo$vitalRates, .x, 'R2.RDS')) |>
      bind_cols(sim = .x)
  ) |>
  left_join(
    sp_info |>
      select(species_id_old, species_name),
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(x = R2reg, y = fct_reorder(species_name, R2reg), fill = sim) +
  geom_density_ridges(alpha = 0.8, scale= 1.8, color = NA) +
  labs(
    x = 'Rsquared (linear regression)',
    y = '',
    fill = 'Model'
  ) +
  theme_classic() +
  theme(axis.text.y = element_text(face = "italic"))
```

<!-- 
# Leave-one-out cross-validation to compare models -->

```{r loo, eval=FALSE, echo = FALSE, warning=FALSE,fig.height = 8, fig.width = 8, warning=FALSE, message=FALSE}
map_dfr(
    spIds,
    ~ map2(
        setNames(sim_toLoad, sim_toLoad),
        .x,
        ~ readRDS(paste0('output/', .x, '/loo_', .y, '.RDS'))
      ) |>
      loo::loo_compare() |>
      as.data.frame() |>
      bind_cols(species_id = .x) |>
      rownames_to_column(var = 'sim')
  ) |>
  select(species_id, sim, elpd_diff, se_diff) |>
  pivot_longer(
    cols = c(elpd_diff, se_diff),
    names_to = 'variable',
    values_to = 'value'
  ) |>
  left_join(
    sp_info |>
      select(species_id_old, species_name),
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(x = value, y = species_id, color = sim) +
  geom_point() +
  facet_wrap(~ variable, scales = 'free_x') +
  labs(
    x = '',
    y = '',
    color = 'Model'
  ) +
  theme_classic() +
  theme(legend.position = 'bottom') +
  theme(axis.text.y = element_text(face = "italic"))
```
