---
 title: "MCMC diagnostics - size ingrowth"
 subtitle: "Model: truncated normal + intercept"
 author: "Will Vieira"
 date: "`r paste('Last updated on', format(Sys.time(), '%d %B, %Y'))`"
---

```{r, echo = F, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(cmdstanr))
suppressPackageStartupMessages(library(loo))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(tidybayes))
suppressPackageStartupMessages(library(bayesplot))
suppressPackageStartupMessages(library(ggridges))
suppressPackageStartupMessages(require(RColorBrewer))
suppressPackageStartupMessages(library(DT))
suppressPackageStartupMessages(library(ggrepel))
```
```{r load simulations, echo = F,warning=FALSE,message=FALSE}
# Load simulation variables
model_dir <- 'model_sizeIngrowth/time_truc'
simInfo <- yaml::read_yaml(file.path(model_dir, '_simulation_info.yml'))
output_dir <- file.path('output_sim_processed', simInfo$vitalRates, simInfo$simName)

sp_info <- read_csv(file.path('..', 'forest-IPM', 'data', 'species_id.csv'))
spIds <- simInfo$spIds
simName <- simInfo$simName

# List output files and set vector names to species_id
posteriorPop_files <- grep(
  'posterior_pop',
  dir(output_dir, full.names = TRUE),
  value = TRUE
)
names(posteriorPop_files) <- names(sort(sapply(spIds, function(x) grep(x, posteriorPop_files))))

diag_files <- grep(
  'diagnostics',
  dir(output_dir, full.names = TRUE),
  value = TRUE
)
names(diag_files) <- names(sort(sapply(spIds, function(x) grep(x, diag_files))))
```


```{r load dataset,echo =FALSE,warning=FALSE,message=FALSE}
dataSource <- readRDS(
  gsub('../../', '', simInfo$dataSources)
) |>
filter(species_id %in% spIds)
```

# Rhat

```{r, echo = F,fig.height = 7, fig.width = 7,warning=FALSE,message=FALSE}
imap_dfr(
    diag_files,
    ~ readRDS(.x)[['rhat']] |>
        bind_cols(species_id = .y)
  ) |>
  left_join(
    sp_info,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot(aes(x = rhat, y = fct_reorder(species_name, rhat))) +
    geom_boxplot() +
    theme_classic() +
    geom_vline(xintercept = 1.01, linetype = 2) +
    ylab('') +
    theme(
      axis.text.y = element_text(face = "italic")
    )
```

# Divergent transitions

```{r divergentTransitions, echo = FALSE,warning=FALSE,message=FALSE}
imap_dfr(
    diag_files,
    ~ readRDS(.x)[['diag_summary']]['num_divergent'] |>
        bind_cols(species_id = .y) |>
        mutate(chain = row_number())
  ) |>
  left_join(
    sp_info |> select(species_id_old, species_name),
    by = c('species_id' = 'species_id_old')
  ) |>
  select(!species_id) |>
  pivot_wider(
    names_from = chain,
    values_from = num_divergent,
    names_prefix = 'chain '
  ) |>
  DT::datatable()
```

# Posterior distribution of parameters

```{r posteriorDist, eval = TRUE,echo = F, fig.height = 7, fig.width = 7,warning=FALSE,message=FALSE}
imap_dfr(
    posteriorPop_files,
    ~ readRDS(.x) |>
      bind_cols(species_id = .y)
  ) |>
  left_join(
    sp_info |> select(species_id_old, species_name),
    by = c('species_id' = 'species_id_old')
  ) ->
post

for(Par in unique(post$par))
{
  print(
    post |>
      filter(par == Par) |>
      ggplot() +
      aes(x = value, y = fct_reorder(species_name, value)) +
      stat_eye() +
      ggtitle(Par) +
      xlab('') +
      theme_classic() +
      ylab('') +
      theme(axis.text.y = element_text(face = "italic"))
  )
}
```

# Correlation between parameters

```{r parCorrelation, eval = TRUE,echo = F, fig.height = 8, fig.width = 8,warning=FALSE,message=FALSE}
parComb <- post |>
  filter(par != 'lp__') |>
  reframe(uqPar = unique(par)) |>
  pull(uqPar) |>
  combn(2)

for(i in 1:ncol(parComb))
{
  print(
    post |>
      filter(par %in% parComb[, i]) |>
      filter(iter %in% sample(iter, 1000)) |>
      select(par, value, iter, species_name) |>
      pivot_wider(names_from = par, values_from = value) |>
      ggplot() +
      aes_string(x = parComb[1, i], y = parComb[2, i]) + 
      geom_point(alpha = .5, size = 0.5) +
      facet_wrap(~ species_name) +
      xlab(parComb[1, i]) +
      ylab(parComb[2, i]) +
      theme_minimal() +
      theme(strip.text = element_text(face = 'italic', size = 7))
  )
}
```

# Predictions

## Size of ingrowth in function of time

```{r,echo=FALSE,fig.height=6,fig.width=8,warning=FALSE,message=FALSE}
post |>
  pivot_wider(
    names_from = par,
    values_from = value
  ) |>
  group_by(species_name) |>
  expand_grid(deltaYear = seq(0, 50, 1)) |>
  mutate(
    ing = size_int + phi_time * deltaYear
  ) ->
pred_data

pred_data |>
  ggplot() +
  aes(deltaYear, ing) +
  aes(color = species_name) +
  stat_lineribbon(.width = 0.95, alpha = .8) +
  scale_fill_brewer() +
  theme_classic() +
  labs(
    x = 'Delta time (years)',
    y = 'Ingrowth size',
    color = ''
  ) +
  geom_text_repel(
    data = pred_data |>
      group_by(species_name) |>
      mutate(
        maxYear = max(deltaYear),
        mort = mean(ing[deltaYear == maxYear])
      ) |>
      slice_head(n = 1),
    mapping = aes(x = maxYear, y = mort, label = species_name),
    max.overlaps = Inf,
    nudge_x = 20, direction = 'y', hjust = 'left',
    fontface = 'italic',
    size = 3
  ) +
  theme(
    legend.position = 'none'
  )
```



# Observed vs predicted

```{r crossValidation, echo=FALSE, warning=FALSE, message=FALSE, fig.height=10,fig.width=9}
R2_ls <- list()

for(Sp in spIds)
{
  val_sp <- dataSource[species_id == Sp]

  # replicate each observation by N_iteration to accommodate the parameters draw
  # Using only 200 iterations of the 8000
  val_sp[, obsID := .I]
  val_sp <- val_sp[rep(obsID, each = 200)]
  sampledIter <- sample(1:(simInfo$maxIter/4 * simInfo$nC), 200)
  val_sp[, iter := sampledIter, by = obsID]

  # merge global parameters
  val_sp[
    post |>
        filter(species_id == Sp) |>
        select(-species_id) |>
        pivot_wider(names_from = par, values_from = value) |>
        as.data.table(),
    `:=`(
      species_name = i.species_name,
      size_int = i.size_int,
      phi_time = i.phi_time,
      sigma_size = i.sigma_size
    ),
    on = 'iter'
  ]

  # random iter were selected, create a new iter vetor to a proprer sequence
  # from 1 to nSample of iteration
  val_sp[, iter2 := 1:.N, by = obsID]

  # compute ingrowh size mean
  val_sp[,
    ing_mean := size_int + phi_time * deltaYear_plot
  ]

  # predict ingrowth size
  val_sp[,
    ing_random := qnorm(runif(.N, pnorm(127, ing_mean, sigma_size)), ing_mean, sigma_size)

  ]

  # Gelman R2
  R2gel <- val_sp |>
    group_by(iter2) |>
    summarise(var_size = var(ing_mean, na.rm = TRUE)) |>
    bind_cols(iter = sampledIter) |>
    left_join(
      post |>
        filter(species_id == Sp & par == 'sigma_size') |>
        mutate(sigma_size = value) |>
        select(iter, sigma_size, species_id)
    ) |>
    mutate(R2gelman = var_size/(var_size + sigma_size^2)) |>
    select(iter, iter2, species_id, R2gelman) |>
    setDT()

  # calculate R2 from regression linear model using the 1:1 obs vs pred
  get_r2 <- function(x, y)
      return(summary(lm(y~x))$r.squared)

  R2_ls[[Sp]] <- merge(
    R2gel,
    val_sp[,
      .(
        R2reg = get_r2(dbh, ing_random)
      ), 
      by = iter2
    ],
    by = 'iter2'
  )

  # plot predicted vs observed
  val_sp |>
    select(obsID, deltaYear_plot, dbh, ing_random) |>
    group_by(obsID) |>
    reframe( 
      deltaYear_plot = mean(deltaYear_plot),
      size_obs = mean(dbh),
      size_pred = mean(ing_random)
    ) |>
    pivot_longer(
      cols = contains('size_')
    ) |>
    mutate(name = gsub('size_', '', name)) |>
    ggplot() +
    aes(factor(deltaYear_plot), value) +
    aes(color = name) +
    geom_boxplot(alpha = .8) +
    labs(
      x = 'Delta time (years)',
      y = 'Ingrowth size (mm)',
      color = ''
    ) +
    theme_classic() ->
  p1

  pred_matrix <- val_sp[,
      .(ing_random, iter2, obsID)
    ] |>
    pivot_wider(names_from = obsID, values_from = ing_random) |>
    select(-iter2) |>
    as.matrix()

  val_sp[,
      .(
        m_obs = mean(dbh),
        m_pred = mean(ing_random)),
      by = obsID
    ] |>
    ggplot() +
    aes(m_obs, m_pred) +
    geom_hex(bins = 100) +
    geom_abline(slope = 1, intercept = 0) +
    theme_classic() +
    theme(
      legend.position = "none"
    ) +
    labs(
      x = 'Observerd growth',
      y = 'Predicted growth'
    ) ->
  p2

  ppc_dens_overlay(
      val_sp[iter2 == 1, dbh],
      pred_matrix
    ) +
    xlab('Obs vs predicted size') +
    ylab('') +
    theme_classic() ->
  p3

  print(
    ggarrange(
      p1,
      ggarrange(
        p2, p3,
        ncol = 2,
        legend = 'bottom'
      ),
      nrow = 2,
      labels = unique(val_sp$species_name),
      font.label = list(size = 13, face = 'italic')
    )
  )
}

# save accuracy for model comparison
invisible(
    map_dfr(
    c('R2'),
    ~ get(paste0(.x, '_ls')) |>
        bind_rows() |>
        saveRDS(file = file.path(
          output_dir,
          paste0(.x, '.RDS')
      )
    )
  )
)
```


# Rsquared

The distribution of R squared values is calculated using the Gelman et al. [2018](http://www.stat.columbia.edu/~gelman/research/unpublished/bayes_R2.pdf) definition.

```{r plot Rsquared_gelman,echo=FALSE,fig.height = 7, fig.width = 7, warning=FALSE, message=FALSE}
sim_toLoad <- c(simInfo$simCompare, simName)

map_dfr(
    sim_toLoad,
    ~ readRDS(file.path('output_sim_processed', simInfo$vitalRates, .x, 'R2.RDS')) |>
      bind_cols(sim = .x)
  ) |>
  left_join(
    sp_info |>
      select(species_id_old, species_name),
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(x = R2gelman, y = fct_reorder(species_name, R2gelman), fill = sim) +
  geom_density_ridges(alpha = 0.8, scale= 1.8, color = NA) +
  labs(
    x = 'Rsquared (Gelman 2018)',
    y = '',
    fill = 'Model'
  ) +
  theme_classic() +
  theme(axis.text.y = element_text(face = "italic"))
```

We also calculated the pseudo $R^2$ from the linear model using the 1:1 plot comparing observed values to the predicted values.

```{r plot Rsquared_regression,echo=FALSE,fig.height = 7, fig.width = 7, warning=FALSE, message=FALSE}
map_dfr(
    sim_toLoad,
    ~ readRDS(file.path('output_sim_processed', simInfo$vitalRates, .x, 'R2.RDS')) |>
      bind_cols(sim = .x)
  ) |>
  left_join(
    sp_info |>
      select(species_id_old, species_name),
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(x = R2reg, y = fct_reorder(species_name, R2reg), fill = sim) +
  geom_density_ridges(alpha = 0.8, scale= 1.8, color = NA) +
  labs(
    x = 'Rsquared (linear regression)',
    y = '',
    fill = 'Model'
  ) +
  theme_classic() +
  theme(axis.text.y = element_text(face = "italic"))
```
