---
 title: "MCMC diagnostics - recruitment"
 subtitle: "Model: intercept + plot random effects + competition + climate"
 author: "Will Vieira"
 date: "`r paste('Last updated on', format(Sys.time(), '%d %B, %Y'))`"
---

```{r, echo = F, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(cmdstanr))
suppressPackageStartupMessages(library(loo))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(tidybayes))
suppressPackageStartupMessages(library(bayesplot))
suppressPackageStartupMessages(library(ggridges))
suppressPackageStartupMessages(require(RColorBrewer))
suppressPackageStartupMessages(library(DT))
suppressPackageStartupMessages(library(ggrepel))
```
```{r load simulations, echo = F,warning=FALSE,message=FALSE}
# Load simulation variables
model_dir <- 'model_recruitment/intcpt_plot_comp_clim'
simInfo <- yaml::read_yaml(file.path(model_dir, '_simulation_info.yml'))
output_dir <- file.path('output_sim_processed', simInfo$vitalRates, simInfo$simName)

sp_info <- read_csv(file.path('..', 'forest-IPM', 'data', 'species_id.csv'))
spIds <- simInfo$spIds
simName <- simInfo$simName

# List output files and set vector names to species_id
posteriorPop_files <- grep(
  'posterior_pop',
  dir(output_dir, full.names = TRUE),
  value = TRUE
)
names(posteriorPop_files) <- names(sort(sapply(spIds, function(x) grep(x, posteriorPop_files))))

posteriormPlot_files <- grep(
  'posterior_plot',
  dir(output_dir, full.names = TRUE),
  value = TRUE
)
names(posteriormPlot_files) <- names(sort(sapply(spIds, function(x) grep(x, posteriormPlot_files))))


diag_files <- grep(
  'diagnostics',
  dir(output_dir, full.names = TRUE),
  value = TRUE
)
names(diag_files) <- names(sort(sapply(spIds, function(x) grep(x, diag_files))))
```


```{r load dataset,echo =FALSE,warning=FALSE,message=FALSE}
dataSource <- readRDS(
  gsub('../../', '', simInfo$dataSources)
)
```

# Rhat

```{r, echo = F,fig.height = 7, fig.width = 7,warning=FALSE,message=FALSE}
imap_dfr(
    diag_files,
    ~ readRDS(.x)[['rhat']] |>
        bind_cols(species_id = .y)
  ) |>
  left_join(
    sp_info,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot(aes(x = rhat, y = fct_reorder(species_name, rhat))) +
    geom_boxplot() +
    theme_classic() +
    geom_vline(xintercept = 1.01, linetype = 2) +
    ylab('') +
    theme(
      axis.text.y = element_text(face = "italic")
    )
```

# Divergent transitions

```{r divergentTransitions, echo = FALSE,warning=FALSE,message=FALSE}
imap_dfr(
    diag_files,
    ~ readRDS(.x)[['diag_summary']]['num_divergent'] |>
        bind_cols(species_id = .y) |>
        mutate(chain = row_number())
  ) |>
  left_join(
    sp_info |> select(species_id_old, species_name),
    by = c('species_id' = 'species_id_old')
  ) |>
  select(!species_id) |>
  pivot_wider(
    names_from = chain,
    values_from = num_divergent,
    names_prefix = 'chain '
  ) |>
  DT::datatable()
```

# Posterior distribution of parameters

```{r posteriorDist, eval = TRUE,echo = F, fig.height = 7, fig.width = 7,warning=FALSE,message=FALSE}
imap_dfr(
    posteriorPop_files,
    ~ readRDS(.x) |>
      bind_cols(species_id = .y)
  ) |>
  left_join(
    sp_info |> select(species_id_old, species_name),
    by = c('species_id' = 'species_id_old')
  ) ->
post

for(Par in unique(post$par))
{
  print(
    post |>
      filter(par == Par) |>
      ggplot() +
      aes(x = value, y = fct_reorder(species_name, value)) +
      stat_eye() +
      ggtitle(Par) +
      xlab('') +
      theme_classic() +
      ylab('') +
      theme(axis.text.y = element_text(face = "italic"))
  )
}
```

# Correlation between parameters

```{r parCorrelation, eval = TRUE,echo = F, fig.height = 8, fig.width = 8,warning=FALSE,message=FALSE}
parComb <- post |>
  filter(par != 'lp__') |>
  reframe(uqPar = unique(par)) |>
  pull(uqPar) |>
  combn(2)

for(i in 1:ncol(parComb))
{
  print(
    post |>
      filter(par %in% parComb[, i]) |>
      filter(iter %in% sample(iter, 1000)) |>
      select(par, value, iter, species_name) |>
      pivot_wider(names_from = par, values_from = value) |>
      ggplot() +
      aes_string(x = parComb[1, i], y = parComb[2, i]) + 
      geom_point(alpha = .5, size = 0.5) +
      facet_wrap(~ species_name) +
      xlab(parComb[1, i]) +
      ylab(parComb[2, i]) +
      theme_minimal() +
      theme(strip.text = element_text(face = 'italic', size = 7))
  )
}
```

# Predictions

## Ingrowth in function of time

```{r,echo=FALSE,fig.height=5,fig.width=8,warning=FALSE,message=FALSE}
post |>
  pivot_wider(
    names_from = par,
    values_from = value
  ) |>
  group_by(species_name) |>
  expand_grid(deltaYear = seq(0, 100, 5)) |>
  mutate(
    lambda = exp(mPop_log) * 399 *
             (1 - exp(-exp(p_log))^deltaYear)/(1 - exp(-exp(p_log)))
  ) ->
pred_data

pred_data |>
  filter(lambda < quantile(lambda, .999)) |>
  ggplot() +
  aes(x = deltaYear, y = lambda, color = species_name) +
  stat_lineribbon(.width = 0.95, alpha = .8) +
  scale_fill_brewer() +
  theme_classic() +
  labs(
    x = 'Delta time (years)',
    y = 'Number of ingrowth recruits',
    color = ''
  ) +
  geom_text_repel(
    data = pred_data |>
      group_by(species_name) |>
      mutate(
        maxYear = max(deltaYear),
        lambda = mean(lambda[deltaYear == maxYear])
      ) |>
      slice_head(n = 1),
    mapping = aes(x = maxYear, y = lambda, label = species_name),
    max.overlaps = Inf,
    nudge_x = 20, direction = 'y', hjust = 'left',
    fontface = 'italic',
    size = 3
  ) +
  theme(
    legend.position = 'none'
  )
```


## Conspecific basal area effect on ingrowth rate `m`

```{r,echo=FALSE,fig.height=5,fig.width=8,warning=FALSE,message=FALSE}
post |>
  filter(par %in% c('mPop_log', 'sigma_BA', 'optimal_BA')) |>
  pivot_wider(
    names_from = par,
    values_from = value
  ) |>
  group_by(species_name) |>
  expand_grid(BA_adult_sp = seq(0, 70, 2.5)) |>
  mutate(
    BAeffect = (-1/sigma_BA^2) * (BA_adult_sp - optimal_BA)^2,
    m = exp(mPop_log + BAeffect)
  ) ->
pred_data
 
pred_data |>
  filter(m < quantile(m, 0.999)) |>
  ggplot() +
  aes(x = BA_adult_sp, y = m, color = species_name) +
  stat_lineribbon(.width = 0.95, alpha = .8) +
  scale_fill_brewer() +
  labs(
    x = 'Conspecific basal area',
    y = ''
  ) +
  geom_text_repel(
    data = pred_data |>
      group_by(species_name) |>
      mutate(
        max_x = max(BA_adult_sp),
        max_y = mean(m[BA_adult_sp == max_x])
      ) |>
      slice_head(n = 1),
    mapping = aes(x = max_x, y = max_y, label = species_name),
    max.overlaps = Inf,
    nudge_x = 10, direction = 'y', hjust = 'left',
    fontface = 'italic',
    size = 3,
    segment.alpha = 0.5
  ) +
  theme_classic() +
  theme(legend.position = 'none')
```

## Plot total basal area effect on survival `p`

```{r,echo=FALSE,fig.height=5,fig.width=8,warning=FALSE,message=FALSE}
post |>
  filter(par %in% c('p_log', 'beta_p')) |>
  pivot_wider(
    names_from = par,
    values_from = value
  ) |>
  group_by(species_name) |>
  expand_grid(BA_adult = seq(0, 70, 3)) |>
  mutate(
    BAeffect = BA_adult * -beta_p,
    p = exp(-exp(p_log) + BAeffect)
  ) ->
pred_data

pred_data |>
  filter(p < quantile(p, 0.999)) |>
  ggplot() +
  aes(x = BA_adult, y = p, color = species_name) +
  stat_lineribbon(.width = 0.95, alpha = .8) +
  scale_fill_brewer() +
  labs(
    x = 'Basal area competition',
    y = ''
  ) +
  geom_text_repel(
    data = pred_data |>
      group_by(species_name) |>
      mutate(
        max_x = max(BA_adult),
        max_y = mean(p[BA_adult == max_x])
      ) |>
      slice_head(n = 1),
    mapping = aes(x = max_x, y = max_y, label = species_name),
    max.overlaps = Inf,
    nudge_x = 10, direction = 'y', hjust = 'left',
    fontface = 'italic',
    size = 3,
    segment.alpha = 0.5
  ) +
  theme_classic() +
  theme(legend.position = 'none')
```


## Mean annual temperature effect on ingrowth rate `m`

```{r,echo=FALSE,fig.height=5,fig.width=8,warning=FALSE,message=FALSE}
post |>
  filter(par %in% c('mPop_log', 'optimal_temp', 'tau_temp')) |>
  pivot_wider(
    names_from = par,
    values_from = value
  ) |>
  group_by(species_name) |>
  expand_grid(temp = seq(0, 1, 0.05)) |>
  mutate(
    TempEffect = -tau_temp * (temp - optimal_temp)^2,
    m = exp(mPop_log + TempEffect)
  ) ->
pred_data
 
pred_data |>
  filter(m < quantile(m, 0.999)) |>
  ggplot() +
  aes(temp, m, color = species_name) +
  stat_lineribbon(.width = 0.95, alpha = .8) +
  scale_fill_brewer() +
  labs(
    x = 'Mean annual temperature',
    y = ''
  ) +
  geom_text_repel(
    data = pred_data |>
      group_by(species_name) |>
      mutate(
        max_x = max(temp),
        max_y = mean(m[temp == max_x])
      ) |>
      slice_head(n = 1),
    mapping = aes(x = max_x, y = max_y, label = species_name),
    max.overlaps = Inf,
    nudge_x = 0.15, direction = 'y', hjust = 'left',
    fontface = 'italic',
    size = 3,
    segment.alpha = 0.5
  ) +
  theme_classic() +
  theme(legend.position = 'none')
```


## Mean annual precipitation effect on ingrowth rate `m`

```{r,echo=FALSE,fig.height=9,fig.width=9,warning=FALSE,message=FALSE}
post |>
  filter(par %in% c('mPop_log', 'optimal_prec', 'tau_prec')) |>
  pivot_wider(
    names_from = par,
    values_from = value
  ) |>
  group_by(species_name) |>
  expand_grid(prec = seq(0, 1, 0.05)) |>
  mutate(
    precEffect = -tau_prec * (prec - optimal_prec)^2,
    m = exp(mPop_log + precEffect)
  ) ->
pred_data
 
pred_data |>
  filter(m < quantile(m, 0.999)) |>
  ggplot() +
  aes(prec, m, color = species_name) +
  stat_lineribbon(.width = 0.95, alpha = .8) +
  scale_fill_brewer() +
  labs(
    x = 'Mean annual precipitation',
    y = ''
  ) +
  geom_text_repel(
    data = pred_data |>
      group_by(species_name) |>
      mutate(
        max_x = max(prec),
        max_y = mean(m[prec == max_x])
      ) |>
      slice_head(n = 1),
    mapping = aes(x = max_x, y = max_y, label = species_name),
    max.overlaps = Inf,
    nudge_x = 0.15, direction = 'y', hjust = 'left',
    fontface = 'italic',
    size = 3,
    segment.alpha = 0.5
  ) +
  theme_classic() +
  theme(legend.position = 'none')
```

# Observed vs predicted

```{r crossValidation, echo=FALSE, warning=FALSE, message=FALSE, fig.height=12,fig.width=10}
# As we are using each draw of the posterior distribution to predict the validation data, the dataset becames too big to store all species in memory.
# So I will loop over each species to (i) load dataset, (ii) predict, (iii) compute prediction metrics and (iv) generate the plots

# list to save MSE and R2 for each species to be used later
MSE_list <- R2_list <- list()

for(Sp in spIds)
{
  val_sp <- dataSource[species_id == Sp]

  # remove NAs
  val_sp <- val_sp[
    !is.na(BA_adult),
    .(
      plot_id, species_id, nbRecruit, deltaYear_plot,
      plot_size, BA_adult, BA_adult_sp, bio_01_mean_scl, bio_12_mean_scl
    )
  ]

  # replicate each observation by N_iteration to accommodate the parameters draw
  # Using only 200 iterations of the 8000
  val_sp[, obsID := .I]
  val_sp <- val_sp[rep(obsID, each = 200)]
  sampledIter <- sample(1:(simInfo$maxIter/4 * simInfo$nC), 200)
  val_sp[, iter := sampledIter, by = obsID]

  # merge global parameters
  val_sp[
    post |>
        filter(species_id == Sp) |>
        select(-species_id) |>
        pivot_wider(names_from = par, values_from = value) |>
        as.data.table(),
    `:=`(
      species_name = i.species_name,
      mPop_log = i.mPop_log,
      p_log = i.p_log,
      sigma_plot = i.sigma_plot,
      optimal_BA = i.optimal_BA,
      sigma_BA = i.sigma_BA,
      beta_p = i.beta_p,
      optimal_temp = i.optimal_temp,
      tau_temp = i.tau_temp,
      optimal_prec = i.optimal_prec,
      tau_prec = i.tau_prec
    ),
    on = 'iter'
  ]

  # merge plot specific parameters
  val_sp[
    readRDS(posteriormPlot_files[Sp]) |>
      setDT(),
    mPlot_log := i.value,
    on = c('iter', 'plot_id')
  ]

  # random iter were selected, create a new iter vetor to a proprer sequence
  # from 1 to nSample of iteration
  val_sp[, iter2 := 1:.N, by = obsID]

  # compute m plot specific parameter with BA_adult_sp effect
  val_sp[,
    m := exp(
      mPop_log + mPlot_log +
      (-1/sigma_BA^2) * (BA_adult_sp - optimal_BA)^2 +
      -tau_temp * (bio_01_mean_scl - optimal_temp)^2 +
      -tau_prec * (bio_12_mean_scl - optimal_prec)^2 
    )
  ]

  # compute p parameter in function of BA_adult
  val_sp[,
    p := exp(
      -exp(
        p_log
      ) +
      BA_adult * -beta_p
    )
  ]

  # predict number of recruit
  val_sp[,
    lambda := m *
              plot_size *
              (1 - p^deltaYear_plot)/(1 - p)
  ]

  # simulate nbRecruit from model distribution (Poisson dist)
  val_sp[, lambda_random := rpois(.N, lambda)]

  # calculate MSE
  val_sp[, SE := (nbRecruit - lambda_random)^2]
  
  # mean SE over all observations for each iteration and save to list
  MSE_list[[Sp]] <- val_sp[, 
    .(
      MSE = mean(SE, na.rm = TRUE),
      species_id = Sp
    ),
    by = iter2
  ]

  # Calculate R2 (Gelman et al 2018)
  R2gel <- val_sp |>
    group_by(iter2) |>
    reframe(
      var_lambda = var(lambda, na.rm = TRUE),
      mean_lambda = mean(lambda, na.rm = TRUE)      
    ) |>
    bind_cols(
      iter = sampledIter,
      species_id = Sp
    ) |>
    mutate(
      R2gelman = var_lambda/(var_lambda + mean_lambda)
    ) |>
    select(iter, iter2, species_id, R2gelman) |>
    setDT()

  # calculate R2 from regression linear model using the 1:1 obs vs pred
  get_r2 <- function(x, y)
      return(summary(lm(y~x))$r.squared)
  
  R2_list[[Sp]] <- merge(
    R2gel,
    val_sp[, .(R2reg = get_r2(nbRecruit, lambda_random)), by = iter2],
    by = 'iter2'
  )


  # parameter mean
  post |>
    filter(species_id == Sp) |>
    group_by(par) |>
    reframe(value = mean(value)) ->
  post_sp_mean
  
  # plot
  p1 <- val_sp[
      nbRecruit < quantile(nbRecruit, 0.999),
      .(BA_adult_sp = unique(BA_adult_sp), m_obs = mean(nbRecruit), m_pred  =mean(lambda_random)),
      by = obsID
    ] |>
    mutate(
      BAsp_cut = cut(BA_adult_sp, seq(0, max(BA_adult_sp), length.out = 10), include.lowest = TRUE)
    ) |>
    select(BAsp_cut, contains('m_')) |>
    pivot_longer(
      cols = !BAsp_cut,
      names_to = 'sim'
    ) |>
    ggplot() +
    aes(BAsp_cut, value, fill = sim) +
    geom_boxplot() +
    labs(
      x = 'Conspecific basal area',
      y = 'Number of recruitment',
      subtitle = unique(val_sp$species_name)
    ) +
    theme_classic() +
    theme(
      legend.position = 'none',
      plot.subtitle = element_text(face = "italic")
    )

  p2 <- val_sp[
      nbRecruit < quantile(nbRecruit, 0.999),
      .(BA_adult = unique(BA_adult), m_obs = mean(nbRecruit), m_pred  =mean(lambda_random)),
      by = obsID
    ] |>
    mutate(
      BA_cut = cut(BA_adult, seq(0, max(BA_adult), length.out = 10), include.lowest = TRUE)
    ) |>
    select(BA_cut, contains('m_')) |>
    pivot_longer(
      cols = !BA_cut,
      names_to = 'sim'
    ) |>
    ggplot() +
    aes(BA_cut, value, fill = sim) +
    geom_boxplot() +
    labs(
      x = 'Total basal area',
      y = 'Number of recruitment',
      fill = '',
      subtitle = ''
    ) +
    theme_classic()

  p3 <- val_sp[
      nbRecruit < quantile(nbRecruit, 0.999),
      .(bio_01_mean_scl = unique(bio_01_mean_scl), m_obs = mean(nbRecruit), m_pred  =mean(lambda_random)),
      by = obsID
    ] |>
    mutate(
      temp_cut = cut(bio_01_mean_scl, seq(min(bio_01_mean_scl), max(bio_01_mean_scl), length.out = 10), include.lowest = TRUE)
    ) |>
    select(temp_cut, contains('m_')) |>
    pivot_longer(
      cols = !temp_cut,
      names_to = 'sim'
    ) |>
    ggplot() +
    aes(temp_cut, value, fill = sim) +
    geom_boxplot() +
    labs(
      x = 'Mean annual temperature',
      y = 'Number of recruitment',
      fill = '',
      subtitle = ''
    ) +
    theme_classic() +
    theme(legend.position = 'none')

  p4 <- val_sp[
      nbRecruit < quantile(nbRecruit, 0.999),
      .(bio_12_mean_scl = unique(bio_12_mean_scl), m_obs = mean(nbRecruit), m_pred = mean(lambda_random)),
      by = obsID
    ] |>
    mutate(
      prec_cut = cut(bio_12_mean_scl, seq(min(bio_12_mean_scl), max(bio_12_mean_scl), length.out = 10), include.lowest = TRUE)
    ) |>
    select(prec_cut, contains('m_')) |>
    pivot_longer(
      cols = !prec_cut,
      names_to = 'sim'
    ) |>
    ggplot() +
    aes(prec_cut, value, fill = sim) +
    geom_boxplot() +
    labs(
      x = 'Mean annual precipitation',
      y = 'Number of recruitment',
      fill = '',
      subtitle = ''
    ) +
    theme_classic()

  # plot
  p5 <- val_sp[
      nbRecruit < quantile(nbRecruit, 0.999),
      .(m_obs=mean(nbRecruit), m_pred=mean(lambda_random)),
      by = obsID
    ] |>
    ggplot() +
    aes(m_obs, m_pred) +
    geom_point(alpha = .3, size = .5) +
    geom_abline(slope = 1, intercept = 0) + 
    geom_smooth(method = 'gam') +
    tune::coord_obs_pred() +
    theme_classic() +
    theme(
      legend.position = 'none',
      plot.margin = margin(t = 30)
    ) +
    xlab('Observerd number of recruitment') +
    ylab('Predicted number of recruitment')

  pred_matrix <- val_sp[, .(lambda_random, iter2, obsID)] |>
      pivot_wider(names_from = obsID, values_from = lambda_random) |>
      select(-iter2) |>
      as.matrix()

  xLim_max <- val_sp[, quantile(nbRecruit, probs = 0.999)]
  p6 <- ppc_dens_overlay(
      val_sp[iter2 == 1, nbRecruit],
      pred_matrix[sample(1:nrow(pred_matrix), 200), ]
  ) +
  xlim(0, xLim_max) +
  xlab('Number of recruitment') +
  theme_classic()

  print(
    ggpubr::ggarrange(
        p1, p2, p3, p4, p5, p6,
        ncol = 2,
        nrow = 3,
        font.label = list(size = 13, face = 'italic')
    ) +
    theme(plot.margin = margin(t = 25))
  )

  rm(p1);rm(p2);rm(val_sp)
}

# save MSE and R2 to be used for model comparison
invisible(
    map_dfr(
    c('MSE', 'R2'),
    ~ get(paste0(.x, '_list')) |>
        bind_rows() |>
        saveRDS(file = file.path(
          output_dir,
          paste0(.x, '.RDS')
      )
    )
  )
)
```


# Mean squared error (MSE)

```{r plot_MSE, echo = FALSE, warning=FALSE, message=FALSE, fig.height = 7, fig.width = 8}
sim_toLoad <- c(simInfo$simCompare, simName)

map_dfr(
    sim_toLoad,
    ~ readRDS(file.path('output_sim_processed', simInfo$vitalRates, .x, 'MSE.RDS')) |>
      bind_cols(sim = .x)
  ) |>
  left_join(
    sp_info |>
      select(species_id_old, species_name),
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(x = MSE, y = fct_reorder(species_name, MSE), fill = sim) +
  geom_density_ridges(alpha = 0.8, scale= 1.8, color = NA) +
  labs(
    x = 'Mean Squared error (MSE)',
    y = '',
    fill = 'Model'
  ) +
  theme_classic() +
  theme(axis.text.y = element_text(face = "italic"))
```


# Rsquared

The distribution of R squared values is calculated using the Gelman et al. [2018](http://www.stat.columbia.edu/~gelman/research/unpublished/bayes_R2.pdf) definition.

```{r plot Rsquared_gelman,echo=FALSE,fig.height = 7, fig.width = 7, warning=FALSE, message=FALSE}
map_dfr(
    sim_toLoad,
    ~ readRDS(file.path('output_sim_processed', simInfo$vitalRates, .x, 'R2.RDS')) |>
      bind_cols(sim = .x)
  ) |>
  left_join(
    sp_info |>
      select(species_id_old, species_name),
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(x = R2gelman, y = fct_reorder(species_name, R2gelman), fill = sim) +
  geom_density_ridges(alpha = 0.8, scale= 1.8, color = NA) +
  labs(
    x = 'Rsquared (Gelman 2018)',
    y = '',
    fill = 'Model'
  ) +
  theme_classic() +
  theme(axis.text.y = element_text(face = "italic"))
```

We also calculated the pseudo $R^2$ from the linear model using the 1:1 plot comparing observed values to the predicted values.

```{r plot Rsquared_regression,echo=FALSE,fig.height = 7, fig.width = 7, warning=FALSE, message=FALSE}
map_dfr(
    sim_toLoad,
    ~ readRDS(file.path('output_sim_processed', simInfo$vitalRates, .x, 'R2.RDS')) |>
      bind_cols(sim = .x)
  ) |>
  left_join(
    sp_info |>
      select(species_id_old, species_name),
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(x = R2reg, y = fct_reorder(species_name, R2reg), fill = sim) +
  geom_density_ridges(alpha = 0.8, scale= 1.8, color = NA) +
  labs(
    x = 'Rsquared (linear regression)',
    y = '',
    fill = 'Model'
  ) +
  theme_classic() +
  theme(axis.text.y = element_text(face = "italic"))
```


# Leave-one-out cross-validation to compare models

```{r loo,echo = FALSE,warning=FALSE,fig.height = 8, fig.width = 8, warning=FALSE, message=FALSE}
map_dfr(
    spIds,
    ~ map2(
        setNames(sim_toLoad, sim_toLoad),
        .x,
        ~ readRDS(paste0('output_sim_processed/', simInfo$vitalRates, '/', .x, '/loo_', .y, '.RDS'))
      ) |>
      loo::loo_compare() |>
      as.data.frame() |>
      bind_cols(species_id = .x) |>
      rownames_to_column(var = 'sim')
  ) |>
  select(species_id, sim, elpd_diff, se_diff) |>
  pivot_longer(
    cols = c(elpd_diff, se_diff),
    names_to = 'variable',
    values_to = 'value'
  ) |>
  left_join(
    sp_info |>
      select(species_id_old, species_name),
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(x = value, y = fct_reorder(species_name, value), color = sim) +
  geom_point() +
  facet_wrap(~ variable, scales = 'free_x') +
  labs(
    x = '',
    y = '',
    color = 'Model'
  ) +
  theme_classic() +
  theme(legend.position = 'bottom') +
  theme(axis.text.y = element_text(face = "italic"))
```


```{r loo2,echo = FALSE,warning=FALSE,fig.height = 4, fig.width = 7, warning=FALSE, message=FALSE}
map_dfr(
    spIds,
    ~ map2(
        setNames(sim_toLoad, sim_toLoad),
        .x,
        ~ readRDS(paste0('output_sim_processed/', simInfo$vitalRates, '/', .x, '/loo_', .y, '.RDS'))
      ) |>
      loo::loo_compare() |>
      as.data.frame() |>
      bind_cols(species_id = .x) |>
      rownames_to_column(var = 'sim')
  ) |>
  select(species_id, sim, elpd_diff, se_diff) |>
  pivot_longer(
    cols = c(elpd_diff, se_diff),
    names_to = 'variable',
    values_to = 'value'
  ) |>
  left_join(
    sp_info |>
      select(species_id_old, species_name),
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(log(value), y = sim) +
  geom_boxplot() +
  facet_wrap(~ variable, scales = 'free_x') +
  theme_classic() +
  ylab('Model')
```
