---
 title: "MCMC diagnostics - survival"
 subtitle: "Model: intercept"
 author: "Will Vieira"
 date: "`r paste('Last updated on', format(Sys.time(), '%d %B, %Y'))`"
---

```{r, echo = F, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(cmdstanr))
suppressPackageStartupMessages(library(loo))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(tidybayes))
suppressPackageStartupMessages(library(bayesplot))
suppressPackageStartupMessages(library(ggridges))
suppressPackageStartupMessages(require(RColorBrewer))
suppressPackageStartupMessages(library(DT))
suppressPackageStartupMessages(library(ggrepel))
```
```{r load simulations, echo = F,warning=FALSE,message=FALSE}
# Load simulation variables
model_dir <- 'model_survival/intcpt'
simInfo <- yaml::read_yaml(file.path(model_dir, '_simulation_info.yml'))
output_dir <- file.path('output_sim_processed', simInfo$vitalRates, simInfo$simName)

sp_info <- read_csv(file.path('..', 'forest-IPM', 'data', 'species_id.csv'))
spIds <- simInfo$spIds
simName <- simInfo$simName

# List output files and set vector names to species_id
posteriorPop_files <- grep(
  'posterior_pop',
  dir(output_dir, full.names = TRUE),
  value = TRUE
)
names(posteriorPop_files) <- names(sort(sapply(spIds, function(x) grep(x, posteriorPop_files))))

# posteriormPlot_files <- grep(
#   'posteriormPlot',
#   dir(foutput_dir, full.names = TRUE),
#   value = TRUE
# )
# names(posteriormPlot_files) <- names(sort(sapply(spIds, function(x) grep(x, posteriormPlot_files))))


diag_files <- grep(
  'diagnostics',
  dir(output_dir, full.names = TRUE),
  value = TRUE
)
names(diag_files) <- names(sort(sapply(spIds, function(x) grep(x, diag_files))))
```


```{r load dataset,echo =FALSE,warning=FALSE,message=FALSE}
dataSource <- readRDS(
  gsub('../../', '', simInfo$dataSources)
)
```

# Rhat

```{r, echo = F,fig.height = 7, fig.width = 7,warning=FALSE,message=FALSE}
imap_dfr(
    diag_files,
    ~ readRDS(.x)[['rhat']] |>
        bind_cols(species_id = .y)
  ) |>
  left_join(
    sp_info,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot(aes(x = rhat, y = fct_reorder(species_name, rhat))) +
    geom_boxplot() +
    theme_classic() +
    geom_vline(xintercept = 1.01, linetype = 2) +
    ylab('') +
    theme(
      axis.text.y = element_text(face = "italic")
    )
```

# Divergent transitions

```{r divergentTransitions, echo = FALSE,warning=FALSE,message=FALSE}
imap_dfr(
    diag_files,
    ~ readRDS(.x)[['diag_summary']]['num_divergent'] |>
        bind_cols(species_id = .y) |>
        mutate(chain = row_number())
  ) |>
  left_join(
    sp_info |> select(species_id_old, species_name),
    by = c('species_id' = 'species_id_old')
  ) |>
  select(!species_id) |>
  pivot_wider(
    names_from = chain,
    values_from = num_divergent,
    names_prefix = 'chain '
  ) |>
  DT::datatable()
```

# Posterior distribution of parameters

```{r posteriorDist, eval = TRUE,echo = F, fig.height = 7, fig.width = 7,warning=FALSE,message=FALSE}
imap_dfr(
    posteriorPop_files,
    ~ readRDS(.x) |>
      bind_cols(species_id = .y)
  ) |>
  left_join(
    sp_info |> select(species_id_old, species_name),
    by = c('species_id' = 'species_id_old')
  ) ->
post

for(Par in unique(post$par))
{
  print(
    post |>
      filter(par == Par) |>
      ggplot() +
      aes(x = value, y = fct_reorder(species_name, value)) +
      stat_eye() +
      ggtitle(Par) +
      xlab('') +
      theme_classic() +
      ylab('') +
      theme(axis.text.y = element_text(face = "italic"))
  )
}
```

# Correlation between parameters

```{r parCorrelation, eval = TRUE,echo = F, fig.height = 8, fig.width = 8,warning=FALSE,message=FALSE}
parComb <- post |>
  # filter(par != 'lp__') |>
  reframe(uqPar = unique(par)) |>
  pull(uqPar) |>
  combn(2)

for(i in 1:ncol(parComb))
{
  print(
    post |>
      filter(par %in% parComb[, i]) |>
      filter(iter %in% sample(iter, 1000)) |>
      select(par, value, iter, species_name) |>
      pivot_wider(names_from = par, values_from = value) |>
      ggplot() +
      aes_string(x = parComb[1, i], y = parComb[2, i]) + 
      geom_point(alpha = .5, size = 0.5) +
      facet_wrap(~ species_name) +
      xlab(parComb[1, i]) +
      ylab(parComb[2, i]) +
      theme_minimal() +
      theme(strip.text = element_text(face = 'italic', size = 7))
  )
}
```

# Predictions

## Mortality probability in function of time

```{r,echo=FALSE,fig.height=6,fig.width=8,warning=FALSE,message=FALSE}
post |>
  pivot_wider(
    names_from = par,
    values_from = value
  ) |>
  group_by(species_name) |>
  expand_grid(deltaYear = seq(0, 1000, 25)) |>
  mutate(
    mort_prob = 1 - (1/(1 + exp(-psi))^deltaYear)
  ) ->
pred_data

pred_data |>
  ggplot() +
  aes(deltaYear, mort_prob) +
  aes(color = species_name) +
  stat_lineribbon(.width = 0.95, alpha = .8) +
  scale_fill_brewer() +
  theme_classic() +
  labs(
    x = 'Delta time (years)',
    y = 'Mortality probability',
    color = ''
  ) +
  geom_text_repel(
    data = pred_data |>
      group_by(species_name) |>
      mutate(
        maxYear = max(deltaYear),
        mort = mean(mort_prob[deltaYear == maxYear])
      ) |>
      slice_head(n = 1),
    mapping = aes(x = maxYear, y = mort, label = species_name),
    max.overlaps = Inf,
    nudge_x = 150, direction = 'y', hjust = 'left',
    fontface = 'italic',
    size = 3
  ) +
  theme(
    legend.position = 'none'
  )
```


# Observed vs predicted

```{r crossValidation, echo=FALSE, warning=FALSE, message=FALSE, fig.height=4,fig.width=9}
accur_ls <- mortProp_ls <- mortPropPlot_ls <- list()

for(Sp in spIds)
{
  val_sp <- dataSource[species_id == Sp]

  # remove NAs
  val_sp <- val_sp[,
    .(
      plot_id, species_id, deltaYear, mort, dbh0,
      BA_comp_sp, BA_comp_inter, bio_01_mean_scl, bio_12_mean_scl
    )
  ]

  # replicate each observation by N_iteration to accommodate the parameters draw
  # Using only 200 iterations of the 8000
  val_sp[, obsID := .I]
  val_sp <- val_sp[rep(obsID, each = 200)]
  sampledIter <- sample(1:(simInfo$maxIter/4 * simInfo$nC), 200)
  val_sp[, iter := sampledIter, by = obsID]

  # merge global parameters
  val_sp[
    post |>
        filter(species_id == Sp) |>
        select(-species_id) |>
        pivot_wider(names_from = par, values_from = value) |>
        as.data.table(),
    `:=`(
      psi = i.psi#,
      # sigma_plot = i.sigma_plot,
      # size_opt = i.size_opt,
      # size_var = i.size_var,
      # Beta = i.Beta,
      # theta = i.theta,
      # optimal_temp = i.optimal_temp,
      # tau_temp = i.tau_temp,
      # optimal_prec = i.optimal_prec,
      # tau_prec = i.tau_prec
    ),
    on = 'iter'
  ]

  # # merge plot specific parameters
  # val_sp[
  #   readRDS(posteriormPlot_files[Sp]) |>
  #     setDT(),
  #   psiPlot := i.value,
  #   on = c('iter', 'plot_id_seq')
  # ]

  # random iter were selected, create a new iter vetor to a proprer sequence
  # from 1 to nSample of iteration
  val_sp[, iter2 := 1:.N, by = obsID]

  # compute longevity logit
  val_sp[,
    longev_log := 1/(1 + exp(
        -(
          psi #+ 
          # psiPlot +
          # -(log(dbh0/size_opt)/size_var)^2 +
          # Beta * (BA_comp_sp + theta * BA_comp_intra) +
          # -tau_temp * (temp_sc - optimal_temp)^2 +
          # -tau_prec * (prec_sc - optimal_prec)^2
        )
      )
    )
  ]

  # compute mortality rate in function of time
  val_sp[,
    mortality_prob := 1 - (longev_log^deltaYear)
  ]

  # predict mortality event
  val_sp[, mort_random := rbinom(.N, size = 1, mortality_prob)]

# Compute model accuracy
  accur_ls[[Sp]] <- val_sp |>
    group_by(iter2) |>
    reframe(
      TP = sum(mort == 1 & mort_random == 1),
      TN = sum(mort == 0 & mort_random == 0),
      FN = sum(mort == 1 & mort_random == 0),
      FP = sum(mort == 0 & mort_random == 1)
    ) |>
    mutate(
      Acc = (TP + TN)/(TP + TN + FP + FN),
      Sensitivity = TP/(TP + FN),
      Specificity = TN/(FP + TN),
      AccCorrected = (Sensitivity + Specificity)/2
    ) |>
    bind_cols(species_id = Sp)

  # Compute proportion of mortality events (pred vs obs)
  mortProp_ls[[Sp]] <- val_sp |>
    group_by(iter2) |>
    reframe(
      mortProp_obs = sum(mort == 1)/n(),
      mortProp_pred = sum(mort_random == 1)/n()
    ) |>
    bind_cols(species_id = Sp)

  mortPropPlot_ls[[Sp]] <- val_sp |>
    group_by(plot_id) |>
    reframe(
      mortProp_obs = sum(mort == 1)/n(),
      mortProp_pred = sum(mort_random == 1)/n()
    ) |>
    bind_cols(species_id = Sp)
}

# save accuracy for model comparison
invisible(
    map_dfr(
    c('accur', 'mortProp', 'mortPropPlot'),
    ~ get(paste0(.x, '_ls')) |>
        bind_rows() |>
        saveRDS(file = file.path(
          output_dir,
          paste0(.x, '.RDS')
      )
    )
  )
)
```

## Accuracy

Using four different metrics:

- **Sensitivity** measures the percentage of dead trees that are identified as dead
- **Specificity** measures the percentage of alive trees that are correctly identified as alive
- **Accuracy** is the ratio of number of correct predictions to the total of predictions
- **Accuracy balanced** considers unbalanced accuracy predictions of positive and negative events (sensitivity and specificity)

```{r plot_accur, echo = FALSE, warning=FALSE, message=FALSE, fig.height = 13, fig.width = 9}
accur_ls |>
  bind_rows() |>
  select(species_id, iter2, Acc, Sensitivity, Specificity, AccCorrected) |>
  pivot_longer(
    cols = c(Acc, Sensitivity, Specificity, AccCorrected)
  ) |>
  mutate(
    name = recode_factor(
      name,
      Acc = 'Accuracy',
      AccCorrected = 'Accuracy corrected',
      Sensitivity = 'Sensitivity',
      Specificity = 'Specificity'
    )
  ) |>
  left_join(
    sp_info,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(value, fct_reorder(species_name, value)) +
  stat_pointinterval() +
  facet_wrap(~ name, scales = 'free_x') +
  theme_classic() +
  labs(
    x = '',
    y = ''
  ) +
  theme(axis.text.y = element_text(face = 'italic'))
```


## Population level mortality rate

```{r plotMortProp,echo=F,warning=FALSE,message=FALSE,fig.height=8,fig.width=7}
propObs <- mortProp_ls |>
  bind_rows() |>
  group_by(species_id) |>
  reframe(mprop = unique(mortProp_obs)) |>
  # bind mean of predicted value so we can sort species by the mean of predict
  # this is necessary to match the sorted species for the next figure
  left_join(
    mortProp_ls |>
      bind_rows() |>
      group_by(species_id) |>
      reframe(mean_pred = mean(mortProp_pred))
  ) |>
  left_join(
    sp_info,
    by = c('species_id' = 'species_id_old')
  ) |>
  arrange(mean_pred)

mortProp_ls |>
  bind_rows() |>
  left_join(
    sp_info,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot() +
  aes(mortProp_pred, y = fct_reorder(species_name, mortProp_pred)) +
  ggridges::geom_density_ridges2(color = NA) +
  geom_segment(
    data = propObs,
    aes(
      x = mprop, xend = mprop,
      y = 1:length(species_name),
      yend = (1:length(species_name)) + .8
    ),
    color = "red", alpha = 0.6
  ) +
  theme_classic() +
  labs(
    x = 'Proportion of mortality events (obs in red vs predicted in grey)',
    y = ''
  ) +
  theme(axis.text.y = element_text(face = 'italic'))
```

```{r plotMortProp2,echo=F,warning=FALSE,message=FALSE,fig.height=10,fig.width=10}
mortPropPlot_ls |>
  bind_rows() |>
  left_join(
    sp_info,
    by = c('species_id' = 'species_id_old')
  ) |>
  ggplot(aes(x = mortProp_obs, y = mortProp_pred)) +
    geom_point(size = .5, alpha = 0.6) +
    facet_wrap(~species_name) +
    geom_smooth(method = 'gam') +
    geom_abline(slope = 1, intercept = 0) +
    theme_minimal() +
    xlab('Observerd mortality rate') +
    ylab('Predicted mortality rate') +
    theme(strip.text = element_text(face = 'italic', size = 7))
```
